---
title: "GAM validation - Bird richness in São Paulo"
author: "Bianca Neves, Irina Lerner, Douglas Cirino"
date: "2024-09-22"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: paper
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)       # For data visualization
library(dplyr)         # For data manipulation
library(PerformanceAnalytics) # For correlation matrix
library(tidyr)         # For data tidying
library(sf)            # For handling spatial data
library(sp)            # For handling spatial data
library(spdep)         # For spatial dependence analysis
library(ggeffects)    # For extracting and plotting model effects
library(mgcv)          # For fitting GAMs
library(viridis)       # For color scales
library(patchwork)    # combine plots
library(car)
library(lme4)
library(MASS)          #nb glm
library(DHARMa)

custom_theme <- theme_minimal() +
  theme(text = element_text(size = 12),
        plot.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

```

# Data 9km²
This is a $9km^2$ grid over the urban area of the city of São Paulo
```{r data, include=FALSE, warning=FALSE}
shp <- st_read("Z:/Bianca/eBird/Data/3km_urb_f5km/richness_effort_v20_total_3km_urb_f5km_30min_10spp_sem_exot_equiv_groups.shp")
#shp <- st_read("C:/Users/Bianca/Documents/USP/Projeto/eBird/Data/1km/richness_effort_v16_total_1km_urb_f1km_30min_10spp.shp")
cdata <- shp


## 2. Rename columns
cdata <- cdata %>% 
  rename(
    richness = richness_s,
    effort = effort_s,
    Bheight_m = Bhght_m,
    Bheight_s = Bhght_s,
    Vheight_m = Vhght_m,
    Vheight_s = Vhght_s,
    Wdistance = Wdistnc,
    contig_f = contg_f,
    contig_h = contg_h,
    pland_f = plant_f,
    pland_h = pland_h,
    area_mn_f = ar_mn_f,
    area_mn_h = ar_mn_h,
    prop_veg = prop_vg,
    prop_for = prp_frs,
    prop_wet = prp__wt,
    prop_grc = prp_grc,
    prop_hrb = prp_hrb,
    landscape_ = lndscp_,
    contig_m_tot = cntg_m_,
    pland_tot = plnd_tt,
    nlsi_tot = nlsi_tt,
    area_m_tot = ar_mn_t,
    Gdistance = Gdistnc,
    geometry = geometry
  )

cdata <- cdata %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  arrange(id)

cdata$centx <- st_coordinates(st_centroid(cdata))[, 1]
cdata$centy <- st_coordinates(st_centroid(cdata))[, 2]

```

## Spatial correlation
```{r spatial, echo=FALSE, warning=FALSE}
# 1. Testing for spatial correlation
coords <- st_centroid(cdata)
nb <- poly2nb(cdata)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
moran.mc(cdata$richness, lw, nsim = 999)
# We see that there is a small spatial correlation

# 2. Check the spatial distribuition of richness
(ggplot(data = cdata) +
  geom_sf(aes(fill = richness), color = NA) +
  scale_fill_viridis_c(direction = -1) +
  custom_theme)
```
There is little evidence of spatial autocorrelation in the richness data, as indicated by the very small Moran's I statistic and the non-significant p-value. This suggests that richness values are not significantly clustered or dispersed across space.


#Variables
## Correlation between variables
```{r cor, echo=FALSE, warning=FALSE}
## 1. Predictors
predictors <- c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")

## 2. Correlation of variables
cor_mat_structure <- as.data.frame(cdata)[, c("richness", predictors)]
suppressWarnings(chart.Correlation(cor_mat_structure, histogram = TRUE, pch = 16, method = "pearson"))
```

## Multicollinearity in the predictors
To check for multicollinearity in the predictors used in the GAM, I calculated the Variance Inflation Factor (VIF) using a linear model (lm). Although the GAM is a more flexible model, the VIF cannot be directly applied to GAMs. Instead, I fit a linear regression model with the same predictors (SEVIo, SUVI, and effort) as in the GAM. The VIF was then calculated from this linear model to assess the degree of multicollinearity between the variables.
```{r vif, echo=FALSE, warning=FALSE}
vif_model <- lm(richness ~ SEVIo + SUVI + effort, data = cdata)
vif(vif_model)
```
VIF values below 5 (or sometimes 10) are generally considered low and indicate that multicollinearity is not a significant concern. For all three predictors (SEVIo, SUVI, and effort), the VIF values are close to 1, which means there is almost no multicollinearity between these variables.

## Association between richness and variables
```{r rich, echo=FALSE, warning=FALSE, message=FALSE}
## Association between richness and variables
# Reshape the data to long format with only 'value' and 'variable'
cdata_long <- cdata[, c("richness", "effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")]
cdata_long <- cdata_long %>% 
  pivot_longer(
    cols = c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb"),
    names_to = "variable",
    values_to = "value"
  )

# Create a faceted plot for the specified predictors
(plot_structure <- ggplot(cdata_long, aes(x = value, y = richness)) + 
    geom_point(alpha = 0.7, color = "black") + 
    geom_smooth(method = "loess", color = "blue", fill = "blue", alpha = 0.1) +
    facet_wrap(~ variable, scales = "free_x", ncol = 2, nrow = 3) +  # Adjusting to 3 rows and 2 columns
    labs(x = "Value", y = "Richness") +
    theme_classic() + 
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 14),
      strip.text = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 16, face = "bold")
    ) +
    ggtitle("Richness  vs. Predictors"))

```

# Generalized Linear Models
## Null model
### GLM 0: Null model
```{r, echo=FALSE, warning=FALSE}
glm_null <- glmer.nb(richness ~ 1 + (1 | effort), 
                     data = cdata)

summary(glm_null)

summary_glm_null <- summary(glm_null)

aic_glm_null <- AIC(glm_null)
print(aic_glm_null)
```

## GLM model
### GLM: rich ~ SEVIo + SGVI
```{r, echo=FALSE, warning=FALSE}
glm_SEVIo_SGVI <- glmer.nb(richness ~ SEVIo + SGVI + (1 | effort), data = cdata)

summary(glm_SEVIo_SGVI)

summary_glm_SEVIo_SGVI <- summary(glm_SEVIo_SGVI)

aic_glm_SEVIo_SGVI <- AIC(glm_SEVIo_SGVI)
print(aic_glm_SEVIo_SGVI)
```

### GLM: rich ~ SEVIo + SUVI
```{r, echo=FALSE, warning=FALSE}
glm_SEVIo_SUVI <- glmer.nb(richness ~ SEVIo + SUVI + (1 | effort), data = cdata)

summary(glm_SEVIo_SUVI)

summary_glm_SEVIo_SUVI <- summary(glm_SEVIo_SUVI)

aic_glm_SEVIo_SUVI <- AIC(glm_SEVIo_SUVI)
print(aic_glm_SEVIo_SUVI)
```

### GLM: rich ~ SEVIo * SGVI
```{r, echo=FALSE, warning=FALSE}
glm_SEVIo_SGVI_int <- glmer.nb(richness ~ SEVIo * SGVI + (1 | effort), data = cdata)

summary(glm_SEVIo_SGVI_int)

summary_glm_SEVIo_SGVI_int <- summary(glm_SEVIo_SGVI_int)

aic_glm_SEVIo_SGVI_int <- AIC(glm_SEVIo_SGVI_int)
print(aic_glm_SEVIo_SGVI_int)
```

### GLM: rich ~ SEVIo * SUVI
```{r, echo=FALSE, warning=FALSE}
glm_SEVIo_SUVI_int <- glmer.nb(richness ~ SEVIo * SUVI + (1 | effort), data = cdata)

summary(glm_SEVIo_SUVI_int)

summary_glm_SEVIo_SUVI_int <- summary(glm_SEVIo_SUVI_int)

aic_glm_SEVIo_SUVI_int <- AIC(glm_SEVIo_SUVI_int)
print(aic_glm_SEVIo_SUVI_int)
```

# Generalized Additive Models
## Null model
### GAM 0: Null model
```{r, echo=FALSE, warning=FALSE}
#### 0. GAM: Null model ####
gam_null <- gam(richness ~ 1 + s(effort, bs = "re"), 
                data = cdata, 
                family = nb(link = "log"))
summary(gam_null)

## 2. AIC
aic_gam_null <- AIC(gam_null)
```

## GAM model
### GAM: rich ~ SEVIo + SGVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo + SGVI ####
## 1. Model
gam_SEVIo_SGVI <- gam(richness ~ s(SEVIo) + s(SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI)

## 2. AIC
aic_gam_SEVIo_SGVI <- AIC(gam_SEVIo_SGVI)
```

```{r, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SGVI
sim_data3 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SGVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SGVI
sim_data3 <- sim_data3 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SGVI, newdata = transform(sim_data3, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_3 <- ggplot(data = sim_data3, aes(x = SEVIo, y = SGVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SGVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of additive model")

plot_heatmap_3
```

### GAM: rich ~ SEVIo + SUVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo + SUVI ####
## 1. Model
gam_SEVIo_SUVI <- gam(richness ~ s(SEVIo) + s(SUVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SUVI)

## 2. AIC
aic_gam_SEVIo_SUVI <- AIC(gam_SEVIo_SUVI)
```

```{r, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SUVI
sim_data3 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SUVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SUVI
sim_data3 <- sim_data3 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SUVI, newdata = transform(sim_data3, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_3 <- ggplot(data = sim_data3, aes(x = SEVIo, y = SUVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SUVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of additive model")

plot_heatmap_3
```


### GAM: rich ~ SEVIo * SGVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo * SGVI ####
## 1. Model
gam_SEVIo_SGVI_int <- gam(richness ~ s(SEVIo,SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI_int)

## 2. AIC
aic_gam_SEVIo_SGVI_int <- AIC(gam_SEVIo_SGVI_int)
```

```{r, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SGVI
sim_data8 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SGVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SGVI
sim_data8 <- sim_data8 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SGVI_int, newdata = transform(sim_data8, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_8 <- ggplot(data = sim_data8, aes(x = SEVIo, y = SGVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SGVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of interaction model")

plot_heatmap_8
```

### GAM: rich ~ SEVIo * SUVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo * SUVI ####
## 1. Model
gam_SEVIo_SUVI_int <- gam(richness ~ s(SEVIo,SUVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SUVI_int)

## 2. AIC
aic_gam_SEVIo_SUVI_int <- AIC(gam_SEVIo_SUVI_int)
```

```{r, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SUVI
sim_data8 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SUVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SUVI
sim_data8 <- sim_data8 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SUVI_int, newdata = transform(sim_data8, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_8 <- ggplot(data = sim_data8, aes(x = SEVIo, y = SUVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SUVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of interaction model")

plot_heatmap_8
```



# Comparing models
## AIC selection
```{r, echo=FALSE, warning=FALSE}
# Store model names and AIC values in a data frame
aic_values <- data.frame(
  Model = c("GLM Model (SEVIo + SGVI)",
            "GLM Model (SEVIo + SUVI)",
            "GLM Model (SEVIo * SGVI)",
            "GLM Model (SEVIo * SUVI)",
            "GAM Model (SEVIo + SGVI)",
            "GAM Model (SEVIo + SUVI)", 
            "GAM Model (SEVIo * SGVI)",
            "GAM Model (SEVIo * SUVI)"),
  AIC = c(aic_glm_SEVIo_SGVI,
          aic_glm_SEVIo_SUVI,
          aic_glm_SEVIo_SGVI_int,
          aic_glm_SEVIo_SUVI_int,
          aic_gam_SEVIo_SGVI, 
          aic_gam_SEVIo_SUVI,
          aic_gam_SEVIo_SGVI_int,
          aic_gam_SEVIo_SUVI_int)
)

# Calculate delta AIC
aic_min <- min(aic_values$AIC)
aic_values$Delta_AIC <- aic_values$AIC - aic_min

# Sort by AIC
aic_values_sorted <- aic_values[order(aic_values$AIC), ]

# Print the sorted AIC values along with Delta AIC
print(aic_values_sorted)
```

# GLM check
Non-normality of the residuals and the heteroscedasticity pattern (non-constant variation of residuals).

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Função para criar o gráfico de resíduos com linha reta em y = 0
create_residual_plot <- function(model, title) {
  residuals_data <- data.frame(
    Fitted = fitted.values(model),
    Residuals = residuals(model)
  )
  
  ggplot(residuals_data, aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.7, color = "black") +
    geom_hline(yintercept = 0, color = "red") +  # Linha reta em y = 0
    labs(x = "Fitted Values", y = "Residuals", title = title) +
    theme_classic() +  
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"), 
      panel.border = element_blank(),  
      axis.line = element_line(color = "black") 
    )
}

# Criando os 4 gráficos de resíduos com títulos ajustados
p1 <- create_residual_plot(glm_SEVIo_SGVI, "GLM SEVIo + SGVI")
p2 <- create_residual_plot(glm_SEVIo_SGVI_int, "GLM SEVIo * SGVI")
p3 <- create_residual_plot(glm_SEVIo_SUVI, "GLM SEVIo + SUVI")
p4 <- create_residual_plot(glm_SEVIo_SUVI_int, "GLM SEVIo * SUVI")

# Organizando os gráficos em um layout 2x2 usando patchwork
layout <- (p1 | p2) / (p3 | p4)
layout

# Função para criar o Q-Q plot com linha de qqline do R base replicada
create_qq_plot <- function(model, title) {
  residuals_data <- data.frame(
    Residuals = sort(residuals(model)),
    Theoretical = sort(qnorm(ppoints(length(residuals(model)))))
  )
  
  # Calcula os quantis para a linha de qqline
  slope <- diff(quantile(residuals_data$Residuals, c(0.25, 0.75))) / 
           diff(quantile(residuals_data$Theoretical, c(0.25, 0.75)))
  intercept <- quantile(residuals_data$Residuals, 0.25) - slope * quantile(residuals_data$Theoretical, 0.25)
  
  ggplot(residuals_data, aes(sample = Residuals)) +
    geom_point(aes(x = Theoretical, y = Residuals), color = "black", alpha = 0.7) +
    geom_abline(intercept = intercept, slope = slope, color = "red") +  # Linha ajustada ao estilo qqline
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles", title = title) +
    theme_classic() +
    theme(
      axis.title = element_text(size = 14), 
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"),  
      panel.border = element_blank(),  
      axis.line = element_line(color = "black")  
    )
}

# Criando os 4 Q-Q plots com títulos ajustados e linha no estilo do R base
qq1 <- create_qq_plot(glm_SEVIo_SGVI, "Q-Q Plot: GLM SEVIo + SGVI")
qq2 <- create_qq_plot(glm_SEVIo_SGVI_int, "Q-Q Plot: GLM SEVIo * SGVI")
qq3 <- create_qq_plot(glm_SEVIo_SUVI, "Q-Q Plot: GLM SEVIo + SUVI")
qq4 <- create_qq_plot(glm_SEVIo_SUVI_int, "Q-Q Plot: GLM SEVIo * SUVI")

# Organizando os Q-Q plots em um layout 2x2 usando patchwork
layout_qq <- (qq1 | qq2) / (qq3 | qq4)
layout_qq
```


```{r, echo=FALSE, warning=FALSE}
# Função para criar o histograma dos resíduos
create_residual_histogram <- function(model, title) {
  residuals_data <- data.frame(Residuals = residuals(model))
  
  ggplot(residuals_data, aes(x = Residuals)) +
    geom_histogram(color = "black", fill = "gray", alpha = 0.6, bins = 30) +  # Histograma com 30 bins
    labs(x = "Residuals", y = "Frequency", title = title) +
    theme_classic() +  
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"), 
      panel.border = element_blank(),  
      axis.line = element_line(color = "black") 
    )
}


# Criando os 4 histogramas dos resíduos
h1 <- create_residual_histogram(glm_SEVIo_SGVI, "Histogram: GLM SEVIo + SGVI")
h2 <- create_residual_histogram(glm_SEVIo_SGVI_int, "Histogram: GLM SEVIo * SGVI")
h3 <- create_residual_histogram(glm_SEVIo_SUVI, "Histogram: GLM SEVIo + SUVI")
h4 <- create_residual_histogram(glm_SEVIo_SUVI_int, "Histogram: GLM SEVIo * SUVI")

# Organizando os gráficos em um layout 2x2 usando patchwork
layout_histograms <- (h1 | h2) / (h3 | h4)
layout_histograms
```
```{r, echo=FALSE, warning=FALSE}
# Teste de normalidade de Shapiro-Wilk
shapiro_test_SEVIo_SGVI <- shapiro.test(residuals(glm_SEVIo_SGVI))
shapiro_test_SEVIo_SGVI_int <- shapiro.test(residuals(glm_SEVIo_SGVI_int))

shapiro_test_SEVIo_SUVI <- shapiro.test(residuals(glm_SEVIo_SUVI))
shapiro_test_SEVIo_SUVI_int <- shapiro.test(residuals(glm_SEVIo_SUVI_int))

# Exibir resultados do teste
print(shapiro_test_SEVIo_SGVI)
print(shapiro_test_SEVIo_SGVI_int)
print(shapiro_test_SEVIo_SUVI)
print(shapiro_test_SEVIo_SUVI_int)

```
```{r, echo=FALSE, warning=FALSE}
# Simular resíduos
sim_resid_SEVIo_SGVI <- simulateResiduals(fittedModel = glm_SEVIo_SGVI)
sim_resid_SEVIo_SGVI_int <- simulateResiduals(fittedModel = glm_SEVIo_SGVI_int)
sim_resid_SEVIo_SUVI <- simulateResiduals(fittedModel = glm_SEVIo_SUVI)
sim_resid_SEVIo_SUVI_int <- simulateResiduals(fittedModel = glm_SEVIo_SUVI_int)

# Plotar resíduos simulados para diagnóstico visual
plot(sim_resid_SEVIo_SGVI)
plot(sim_resid_SEVIo_SGVI_int)
plot(sim_resid_SEVIo_SUVI)
plot(sim_resid_SEVIo_SUVI_int)
```

# GAM check
## GAM: rich ~ SEVIo + SGVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo + SGVI ####
## 1. Model
gam_SEVIo_SGVI <- gam(richness ~ s(SEVIo) + s(SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI)

## 2. AIC
aic_gam_SEVIo_SGVI <- AIC(gam_SEVIo_SGVI)

## 3. Plot
plot(gam_SEVIo_SGVI, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_3 <- vis.gam(gam_SEVIo_SGVI, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SEVIo_SGVI)
gam.check(gam_SEVIo_SGVI)
```

## GAM: rich ~ SEVIo * SGVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo * SGVI ####
## 1. Model
gam_SEVIo_SGVI_int <- gam(richness ~ s(SEVIo,SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI_int)

## 2. AIC
aic_gam_SEVIo_SGVI_int <- AIC(gam_SEVIo_SGVI_int)

## 3. Plot
plot(gam_SEVIo_SGVI_int, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_8 <- vis.gam(gam_SEVIo_SGVI_int, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SEVIo_SGVI_int)
gam.check(gam_SEVIo_SGVI_int)


```

## GAM: rich ~ SEVIo + SUVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo + SUVI ####
## 1. Model
gam_SEVIo_SUVI <- gam(richness ~ s(SEVIo) + s(SUVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SUVI)

## 2. AIC
aic_gam_SEVIo_SUVI <- AIC(gam_SEVIo_SUVI)

## 3. Plot
plot(gam_SEVIo_SUVI, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_3 <- vis.gam(gam_SEVIo_SUVI, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SEVIo_SUVI)
gam.check(gam_SEVIo_SUVI)
```

## GAM: rich ~ SEVIo * SUVI
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo * SUVI ####
## 1. Model
gam_SEVIo_SUVI_int <- gam(richness ~ s(SEVIo,SUVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SUVI_int)

## 2. AIC
aic_gam_SEVIo_SUVI_int <- AIC(gam_SEVIo_SUVI_int)

## 3. Plot
plot(gam_SEVIo_SUVI_int, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_8 <- vis.gam(gam_SEVIo_SUVI_int, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SEVIo_SUVI_int)
gam.check(gam_SEVIo_SUVI_int)


```


## GAM ANOVA
```{r, echo=FALSE, warning=FALSE}
anova(gam_SEVIo_SGVI, gam_SEVIo_SGVI_int, test = "F")
anova(gam_SEVIo_SUVI, gam_SEVIo_SUVI_int, test = "F")
```

# Visualize
```{r, echo=FALSE, warning=FALSE}
ggplot(cdata, aes(x = SGVI, y = richness, color = SEVIo)) +
  geom_point(alpha = 0.7, size=2) + 
  geom_smooth(method = "gam", formula = y ~ s(x), color = "black", fill = "black", alpha = 0.05, size = 0.5) +
  labs(x = "Sparing green", y = "Richness", title = "Richness with sparing green") +
  theme_classic() +  
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12), 
    plot.title = element_text(size = 16, face = "bold"), 
    panel.border = element_blank(),  
    axis.line = element_line(color = "black") 
  )

ggplot(cdata, aes(x = SUVI, y = richness, color = SEVIo)) +
  geom_point(alpha = 0.7, size=2) + 
  geom_smooth(method = "gam", formula = y ~ s(x), color = "black", fill = "black", alpha = 0.05, size = 0.5) +
  labs(x = "Sparing urban", y = "Richness", title = "Richness with sparing urban") +
  theme_classic() +  
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12), 
    plot.title = element_text(size = 16, face = "bold"), 
    panel.border = element_blank(),  
    axis.line = element_line(color = "black") 
  )
```


