---
title: "GAM validation - Bird richness in São Paulo"
author: "Bianca Neves, Irina Lerner, Douglas Cirino"
date: "2024-09-22"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: paper
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)       # For data visualization
library(dplyr)         # For data manipulation
library(PerformanceAnalytics) # For correlation matrix
library(tidyr)         # For data tidying
library(sf)            # For handling spatial data
library(sp)            # For handling spatial data
library(spdep)         # For spatial dependence analysis
library(ggeffects)    # For extracting and plotting model effects
library(mgcv)          # For fitting GAMs
library(viridis)       # For color scales
library(patchwork)    # combine plots
library(car)
library(lme4)
library(MASS)          #nb glm
library(DHARMa)

custom_theme <- theme_minimal() +
  theme(text = element_text(size = 12),
        plot.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

```

# Data 1km²
This is a $1km^2$ grid over the urban area of the city of São Paulo
```{r data, include=FALSE, warning=FALSE}
shp <- st_read("Z:/Bianca/eBird/Data/1km_urb_f1km/richness_effort_v20_total_1km_urb_f1km_30min_10spp_sem_exot_equiv_groups.shp")
#shp <- st_read("C:/Users/Bianca/Documents/USP/Projeto/eBird/Data/1km/richness_effort_v16_total_1km_urb_f1km_30min_10spp.shp")
cdata <- shp


## 2. Rename columns
cdata <- cdata %>% 
  rename(
    richness = richness_s,
    effort = effort_s,
    Bheight_m = Bhght_m,
    Bheight_s = Bhght_s,
    Vheight_m = Vhght_m,
    Vheight_s = Vhght_s,
    Wdistance = Wdistnc,
    contig_f = contg_f,
    contig_h = contg_h,
    pland_f = plant_f,
    pland_h = pland_h,
    area_mn_f = ar_mn_f,
    area_mn_h = ar_mn_h,
    prop_veg = prop_vg,
    prop_for = prp_frs,
    prop_wet = prp_wtl,
    prop_grc = prp_grc,
    prop_hrb = prp_hrb,
    landscape_ = lndscp_,
    contig_m_tot = cntg_m_,
    pland_tot = plnd_tt,
    nlsi_tot = nlsi_tt,
    area_m_tot = ar_mn_t,
    Gdistance = Gdistnc,
    geometry = geometry
  )

cdata <- cdata %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  arrange(id)

cdata$centx <- st_coordinates(st_centroid(cdata))[, 1]
cdata$centy <- st_coordinates(st_centroid(cdata))[, 2]

```

## Spatial correlation
```{r spatial, echo=FALSE, warning=FALSE}
# 1. Testing for spatial correlation
coords <- st_centroid(cdata)
nb <- poly2nb(cdata)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
moran.mc(cdata$richness, lw, nsim = 999)
# We see that there is a small spatial correlation

# 2. Check the spatial distribuition of richness
(ggplot(data = cdata) +
  geom_sf(aes(fill = richness), color = NA) +
  scale_fill_viridis_c(direction = -1) +
  custom_theme)
```

While the Moran's I value of 0.12436 may not seem exceptionally high at first glance, it is considered significant in spatial autocorrelation studies, indicating a clear and notable pattern of clustering in bird richness. 


#Variables
## Correlation between variables
```{r cor, echo=FALSE, warning=FALSE}
## 1. Predictors
predictors <- c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")

## 2. Correlation of variables
cor_mat_structure <- as.data.frame(cdata)[, c("richness", predictors)]
suppressWarnings(chart.Correlation(cor_mat_structure, histogram = TRUE, pch = 16, method = "pearson"))
```

## Multicollinearity in the predictors
To check for multicollinearity in the predictors used in the GAM, I calculated the Variance Inflation Factor (VIF) using a linear model (lm). Although the GAM is a more flexible model, the VIF cannot be directly applied to GAMs. Instead, I fit a linear regression model with the same predictors (SEVIo, SGVI, and effort) as in the GAM. The VIF was then calculated from this linear model to assess the degree of multicollinearity between the variables.
```{r vif, echo=FALSE, warning=FALSE}
vif_model <- lm(richness ~ SEVIo + SGVI + effort, data = cdata)
vif(vif_model)
```
The VIF results confirm that the predictor variables do not exhibit problematic multicollinearity, ensuring the robustness of the model estimates.

## Association between richness and variables
```{r rich, echo=FALSE, warning=FALSE, message=FALSE}
## Association between richness and variables
# Reshape the data to long format with only 'value' and 'variable'
cdata_long <- cdata[, c("richness", "effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")]
cdata_long <- cdata_long %>% 
  pivot_longer(
    cols = c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb"),
    names_to = "variable",
    values_to = "value"
  )


# Create a faceted plot for the specified predictors
(plot_structure <- ggplot(cdata_long, aes(x = value, y = richness)) + 
    geom_point(alpha = 0.7, color = "black") + 
    geom_smooth(method = "loess", color = "blue", fill = "blue", alpha = 0.1) +
    facet_wrap(~ variable, scales = "free_x", ncol = 2, nrow = 3) +  # Adjusting to 3 rows and 2 columns
    labs(x = "Value", y = "Richness") +
    theme_classic() + 
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 14),
      strip.text = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 16, face = "bold")
    ) +
    ggtitle("Richness  vs. Predictors"))

```

# Generalized Linear Models
## Null model
### GLM 0: Null model
```{r, echo=FALSE, warning=FALSE}
glm_null <- glmer.nb(richness ~ 1 + (1 | effort), 
                     data = cdata)

summary(glm_null)

summary_glm_null <- summary(glm_null)

aic_glm_null <- AIC(glm_null)
print(aic_glm_null)
```

## GLM model
### GLM: rich ~ SEVIo + SGVI
```{r, echo=FALSE, warning=FALSE}
glm_SEVIo_SGVI <- glmer.nb(richness ~ SEVIo + SGVI + (1 | effort), data = cdata)

summary(glm_SEVIo_SGVI)

summary_glm_SEVIo_SGVI <- summary(glm_SEVIo_SGVI)

aic_glm_SEVIo_SGVI <- AIC(glm_SEVIo_SGVI)
print(aic_glm_SEVIo_SGVI)
```
```{r, echo=FALSE, warning=FALSE}
glm_SEVIo_SGVI_q <- glmer.nb(richness ~ I(SEVIo^2) + SGVI + (1 | effort), data = cdata)

summary(glm_SEVIo_SGVI_q)

summary_glm_SEVIo_SGVI_q <- summary(glm_SEVIo_SGVI_q)

aic_glm_SEVIo_SGVI_q <- AIC(glm_SEVIo_SGVI_q)
print(aic_glm_SEVIo_SGVI_q)
```


### GLM: rich ~ SEVIo * SGVI
```{r glmint, echo=FALSE, warning=FALSE}
glm_SEVIo_SGVI_int <- glmer.nb(richness ~ SEVIo * SGVI + (1 | effort), data = cdata)

summary(glm_SEVIo_SGVI_int)

summary_glm_SEVIo_SGVI_int <- summary(glm_SEVIo_SGVI_int)

aic_glm_SEVIo_SGVI_int <- AIC(glm_SEVIo_SGVI_int)
print(aic_glm_SEVIo_SGVI_int)
```

# Generalized Additive Models
## Null model
### GAM 0: Null model
```{r gam0, echo=FALSE, warning=FALSE}
#### 0. GAM: Null model ####
gam_null <- gam(richness ~ 1 + s(effort, bs = "re"), 
                data = cdata, 
                family = nb(link = "log"))
summary(gam_null)

## 2. AIC
aic_gam_null <- AIC(gam_null)
```

## GAM model
### GAM: rich ~ SEVIo + SGVI
```{r gam3, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo + SGVI ####
## 1. Model
gam_SEVIo_SGVI <- gam(richness ~ s(SEVIo) + s(SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI)

## 2. AIC
aic_gam_SEVIo_SGVI <- AIC(gam_SEVIo_SGVI)
```

```{r gam3p, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SGVI
sim_data3 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SGVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SGVI
sim_data3 <- sim_data3 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SGVI, newdata = transform(sim_data3, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_3 <- ggplot(data = sim_data3, aes(x = SEVIo, y = SGVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SGVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of additive model")

plot_heatmap_3
```

### GAM: rich ~ SEVIo * SGVI
```{r gam8, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo * SGVI ####
## 1. Model
gam_SEVIo_SGVI_int <- gam(richness ~ s(SEVIo,SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI_int)

## 2. AIC
aic_gam_SEVIo_SGVI_int <- AIC(gam_SEVIo_SGVI_int)
```

```{r gam8p, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SGVI
sim_data8 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SGVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SGVI
sim_data8 <- sim_data8 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SGVI_int, newdata = transform(sim_data8, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_8 <- ggplot(data = sim_data8, aes(x = SEVIo, y = SGVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SGVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of interaction model")

plot_heatmap_8
```

# Comparing models
## AIC selection
```{r compar, echo=FALSE, warning=FALSE}
# Store model names and AIC values in a data frame
aic_values <- data.frame(
  Model = c("GLM Model (SEVIo + SGVI)",
            "GLM Model (I(SEVIo^2) + SGVI)",
            "GLM Model (SEVIo * SGVI)", 
            "GAM Model (SEVIo + SGVI)", 
            "GAM Model (SEVIo * SGVI)"),
  AIC = c(aic_glm_SEVIo_SGVI,
          aic_glm_SEVIo_SGVI_q,
          aic_glm_SEVIo_SGVI_int,
          aic_gam_SEVIo_SGVI, 
          aic_gam_SEVIo_SGVI_int)
)

# Calculate delta AIC
aic_min <- min(aic_values$AIC)
aic_values$Delta_AIC <- aic_values$AIC - aic_min

# Sort by AIC
aic_values_sorted <- aic_values[order(aic_values$AIC), ]

# Print the sorted AIC values along with Delta AIC
print(aic_values_sorted)
```

# GLM check
Non-normality of the residuals and the heteroscedasticity pattern (non-constant variation of residuals).

```{r, echo=FALSE, warning=FALSE}
# Função para criar o gráfico de resíduos com linha reta em y = 0
create_residual_plot <- function(model, title) {
  residuals_data <- data.frame(
    Fitted = fitted.values(model),
    Residuals = residuals(model)
  )
  
  ggplot(residuals_data, aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.7, color = "black") +
    geom_hline(yintercept = 0, color = "red") +  # Linha reta em y = 0
    labs(x = "Fitted Values", y = "Residuals", title = title) +
    theme_classic() +  
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"), 
      panel.border = element_blank(),  
      axis.line = element_line(color = "black") 
    )
}

# Criando os 4 gráficos de resíduos com títulos ajustados
p1 <- create_residual_plot(glm_SEVIo_SGVI, "GLM SEVIo + SGVI")
p2 <- create_residual_plot(glm_SEVIo_SGVI_int, "GLM SEVIo * SGVI")

# Organizando os gráficos em um layout 2x2 usando patchwork
layout <- (p1 | p2)
layout

# Função para criar o Q-Q plot com linha de qqline do R base replicada
create_qq_plot <- function(model, title) {
  residuals_data <- data.frame(
    Residuals = sort(residuals(model)),
    Theoretical = sort(qnorm(ppoints(length(residuals(model)))))
  )
  
  # Calcula os quantis para a linha de qqline
  slope <- diff(quantile(residuals_data$Residuals, c(0.25, 0.75))) / 
           diff(quantile(residuals_data$Theoretical, c(0.25, 0.75)))
  intercept <- quantile(residuals_data$Residuals, 0.25) - slope * quantile(residuals_data$Theoretical, 0.25)
  
  ggplot(residuals_data, aes(sample = Residuals)) +
    geom_point(aes(x = Theoretical, y = Residuals), color = "black", alpha = 0.7) +
    geom_abline(intercept = intercept, slope = slope, color = "red") +  # Linha ajustada ao estilo qqline
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles", title = title) +
    theme_classic() +
    theme(
      axis.title = element_text(size = 14), 
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"),  
      panel.border = element_blank(),  
      axis.line = element_line(color = "black")  
    )
}

# Criando os 4 Q-Q plots com títulos ajustados e linha no estilo do R base
qq1 <- create_qq_plot(glm_SEVIo_SGVI, "Q-Q Plot: GLM SEVIo + SGVI")
qq2 <- create_qq_plot(glm_SEVIo_SGVI_int, "Q-Q Plot: GLM SEVIo * SGVI")


# Organizando os Q-Q plots em um layout 2x2 usando patchwork
layout_qq <- (qq1 | qq2) 
layout_qq
```

```{r, echo=FALSE, warning=FALSE}
# Função para criar o histograma dos resíduos
create_residual_histogram <- function(model, title) {
  residuals_data <- data.frame(Residuals = residuals(model))
  
  ggplot(residuals_data, aes(x = Residuals)) +
    geom_histogram(color = "black", fill = "gray", alpha = 0.6, bins = 30) +  # Histograma com 30 bins
    labs(x = "Residuals", y = "Frequency", title = title) +
    theme_classic() +  
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"), 
      panel.border = element_blank(),  
      axis.line = element_line(color = "black") 
    )
}


# Criando os 4 histogramas dos resíduos
h1 <- create_residual_histogram(glm_SEVIo_SGVI, "Histogram: GLM SEVIo + SGVI")
h2 <- create_residual_histogram(glm_SEVIo_SGVI_int, "Histogram: GLM SEVIo * SGVI")

# Organizando os gráficos em um layout 2x2 usando patchwork
layout_histograms <- (h1 | h2)
layout_histograms
```

```{r, echo=FALSE, warning=FALSE}
# Teste de normalidade de Shapiro-Wilk
shapiro_test_SEVIo_SGVI <- shapiro.test(residuals(glm_SEVIo_SGVI))
shapiro_test_SEVIo_SGVI_int <- shapiro.test(residuals(glm_SEVIo_SGVI_int))

# Exibir resultados do teste
print(shapiro_test_SEVIo_SGVI)
print(shapiro_test_SEVIo_SGVI_int)
```

```{r, echo=FALSE, warning=FALSE}
# Simular resíduos
sim_resid_SEVIo_SGVI <- simulateResiduals(fittedModel = glm_SEVIo_SGVI)
sim_resid_SEVIo_SGVI_int <- simulateResiduals(fittedModel = glm_SEVIo_SGVI_int)

# Plotar resíduos simulados para diagnóstico visual
plot(sim_resid_SEVIo_SGVI)
plot(sim_resid_SEVIo_SGVI_int)
```

# GAM check
## GAM: rich ~ SEVIo + SGVI
The smoothed term for SEVIo has an edf of 3.3087, indicating a moderate nonlinear relationship between SEVIo and species richness. 
The smoothed term for SGVI has an edf of 1.0012, suggesting that its effect is practically linear. 
The value R-sq.(adj) = -0.166 remains negative, which suggests that the complexity of the model may not yet be capturing all the variation in the data.
s(SEVIo): The value of k' = 9 and edf = 3.309 suggest that the choice of k is adequate, with a k-index of 0.98, which is as expected. 
s(SGVI): The k' = 9 and edf = 1.001 also indicate an appropriate fit, but the k-index of 0.90 and p-value of 0.095 suggest that there may be a slight underestimation of the complexity for SGVI. This could be adjusted with a higher value of k, but since the edf is close to 1, linearity must be maintained.
```{r gam3check, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo + SGVI ####
## 1. Model
gam_SEVIo_SGVI <- gam(richness ~ s(SEVIo) + s(SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI)

## 2. AIC
aic_gam_SEVIo_SGVI <- AIC(gam_SEVIo_SGVI)

## 3. Plot
plot(gam_SEVIo_SGVI, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_3 <- vis.gam(gam_SEVIo_SGVI, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SEVIo_SGVI)
gam.check(gam_SEVIo_SGVI)
```

## GAM: rich ~ SEVIo * SGVI
This model indicates that effort and the combination of SEVIo and SGVI are significant predictors of species richness. 
The bivariate smoothed term of SEVIo and SGVI is non-linear and complex, while effort has a significant linear effect.
The value of R-sq.(adj) = -0.173 indicates that the adjusted model, compared to a simple linear model, may not be explaining as much variation as expected. The use of smoothing may be contributing to this result, or there may be important variables missing.
For s(SEVIo, SGVI), the value of k' = 29 and edf = 5.733 indicate that the choice of k is appropriate, as the k-index is 1.04, which is close to the expected value (k-index close to 1) .

```{r gam8check, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo * SGVI ####
## 1. Model
gam_SEVIo_SGVI_int <- gam(richness ~ s(SEVIo,SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SEVIo_SGVI_int)

## 2. AIC
aic_gam_SEVIo_SGVI_int <- AIC(gam_SEVIo_SGVI_int)

## 3. Plot
plot(gam_SEVIo_SGVI_int, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_8 <- vis.gam(gam_SEVIo_SGVI_int, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SEVIo_SGVI_int)
gam.check(gam_SEVIo_SGVI_int)


```

## GAM ANOVA
The data do not show a significant difference in the predictive ability of the two models.
```{r gamanova, echo=FALSE, warning=FALSE}
anova(gam_SEVIo_SGVI, gam_SEVIo_SGVI_int, test = "F")
```

# Visualize
```{r, echo=FALSE, warning=FALSE}
ggplot(cdata, aes(x = SGVI, y = richness, color = SEVIo)) +
  geom_point(alpha = 0.7, size=2) + 
  geom_smooth(method = "gam", formula = y ~ s(x), color = "black", fill = "black", alpha = 0.05, size = 0.5) +
  labs(x = "Sparing green", y = "Richness", title = "Richness with sparing green") +
  theme_classic() +  
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12), 
    plot.title = element_text(size = 16, face = "bold"), 
    panel.border = element_blank(),  
    axis.line = element_line(color = "black") 
  )
```