---
title: "GAM validation - Bird richness in São Paulo"
author: "Bianca Neves, Irina Lerner, Douglas Cirino"
date: "2024-09-22"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: paper
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)       # For data visualization
library(dplyr)         # For data manipulation
library(PerformanceAnalytics) # For correlation matrix
library(tidyr)         # For data tidying
library(sf)            # For handling spatial data
library(sp)            # For handling spatial data
library(spdep)         # For spatial dependence analysis
library(ggeffects)    # For extracting and plotting model effects
library(mgcv)          # For fitting GAMs
library(viridis)       # For color scales
library(patchwork)    # combine plots
library(car)
library(lme4)
library(MASS)          #nb glm
library(DHARMa)

custom_theme <- theme_minimal() +
  theme(text = element_text(size = 12),
        plot.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

```

# Data 4km²
This is a $4km^2$ grid over the urban area of the city of São Paulo
```{r data, include=FALSE, warning=FALSE}
shp <- st_read("Z:/Bianca/eBird/Data/2km_urb_f2.5km/richness_effort_v20_total_2km_urb_f2.5km_30min_10spp_sem_exot_equiv_groups.shp")
#shp <- st_read("C:/Users/Bianca/Documents/USP/Projeto/eBird/Data/1km/richness_effort_v16_total_1km_urb_f1km_30min_10spp.shp")
cdata <- shp


## 2. Rename columns
cdata <- cdata %>% 
  rename(
    richness = richness_s,
    effort = effort_s,
    Bheight_m = Bhght_m,
    Bheight_s = Bhght_s,
    Vheight_m = Vhght_m,
    Vheight_s = Vhght_s,
    Wdistance = Wdistnc,
    contig_f = contg_f,
    contig_h = contg_h,
    pland_f = plant_f,
    pland_h = pland_h,
    area_mn_f = ar_mn_f,
    area_mn_h = ar_mn_h,
    prop_veg = prop_vg,
    prop_for = prp_frs,
    prop_wet = prp__wt,
    prop_grc = prp_grc,
    prop_hrb = prp_hrb,
    landscape_ = lndscp_,
    contig_m_tot = cntg_m_,
    pland_tot = plnd_tt,
    nlsi_tot = nlsi_tt,
    area_m_tot = ar_mn_t,
    Gdistance = Gdistnc,
    geometry = geometry
  )

cdata <- cdata %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  arrange(id)

cdata$centx <- st_coordinates(st_centroid(cdata))[, 1]
cdata$centy <- st_coordinates(st_centroid(cdata))[, 2]

```

## Spatial correlation
```{r spatial, echo=FALSE, warning=FALSE}
# 1. Testing for spatial correlation
coords <- st_centroid(cdata)
nb <- poly2nb(cdata)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
moran.mc(cdata$richness, lw, nsim = 999)
# We see that there is a small spatial correlation

# 2. Check the spatial distribuition of richness
(ggplot(data = cdata) +
  geom_sf(aes(fill = richness), color = NA) +
  scale_fill_viridis_c(direction = -1) +
  custom_theme)
```

While the Moran's I value of 0.18297 may not seem exceptionally high at first glance, it is considered significant in spatial autocorrelation studies, indicating a clear and notable pattern of clustering in bird richness. 


#Variables
## Correlation between variables
```{r cor, echo=FALSE, warning=FALSE}
## 1. Predictors
predictors <- c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")

## 2. Correlation of variables
cor_mat_structure <- as.data.frame(cdata)[, c("richness", predictors)]
suppressWarnings(chart.Correlation(cor_mat_structure, histogram = TRUE, pch = 16, method = "pearson"))
```

## Multicollinearity in the predictors
To check for multicollinearity in the predictors used in the GAM, I calculated the Variance Inflation Factor (VIF) using a linear model (lm). Although the GAM is a more flexible model, the VIF cannot be directly applied to GAMs. Instead, I fit a linear regression model with the same predictors (SVI, prop_arb, and effort) as in the GAM. The VIF was then calculated from this linear model to assess the degree of multicollinearity between the variables.
```{r vif, echo=FALSE, warning=FALSE}
vif_model <- lm(richness ~ SVI + prop_arb + effort, data = cdata)
vif(vif_model)
```
The VIF results confirm that the predictor variables do not exhibit problematic multicollinearity, ensuring the robustness of the model estimates.

## Association between richness and variables
```{r rich, echo=FALSE, warning=FALSE, message=FALSE}
## Association between richness and variables
# Reshape the data to long format with only 'value' and 'variable'
cdata_long <- cdata[, c("richness", "effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")]
cdata_long <- cdata_long %>% 
  pivot_longer(
    cols = c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb"),
    names_to = "variable",
    values_to = "value"
  )

# Create a faceted plot for the specified predictors
(plot_structure <- ggplot(cdata_long, aes(x = value, y = richness)) + 
    geom_point(alpha = 0.7, color = "black") + 
    geom_smooth(method = "loess", color = "blue", fill = "blue", alpha = 0.1) +
    facet_wrap(~ variable, scales = "free_x", ncol = 2, nrow = 3) +  # Adjusting to 3 rows and 2 columns
    labs(x = "Value", y = "Richness") +
    theme_classic() + 
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 14),
      strip.text = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 16, face = "bold")
    ) +
    ggtitle("Richness  vs. Predictors"))

```

# Generalized Linear Models
## Null model
### GLM 0: Null model
```{r, echo=FALSE, warning=FALSE}
glm_null <- glmer.nb(richness ~ 1 + (1 | effort), 
                     data = cdata)

summary(glm_null)

summary_glm_null <- summary(glm_null)

aic_glm_null <- AIC(glm_null)
print(aic_glm_null)
```

## GLM model
### GLM: rich ~ SVI + prop_arb
```{r, echo=FALSE, warning=FALSE}
glm_SVI_prop_arb <- glmer.nb(richness ~ SVI + prop_arb + (1 | effort), data = cdata)

summary(glm_SVI_prop_arb)

summary_glm_SVI_prop_arb <- summary(glm_SVI_prop_arb)

aic_glm_SVI_prop_arb <- AIC(glm_SVI_prop_arb)
print(aic_glm_SVI_prop_arb)
```

### GLM: rich ~ SVI * prop_arb
```{r, echo=FALSE, warning=FALSE}
glm_SVI_prop_arb_int <- glmer.nb(richness ~ SVI * prop_arb + (1 | effort), data = cdata)

summary(glm_SVI_prop_arb_int)

summary_glm_SVI_prop_arb_int <- summary(glm_SVI_prop_arb_int)

aic_glm_SVI_prop_arb_int <- AIC(glm_SVI_prop_arb_int)
print(aic_glm_SVI_prop_arb_int)
```

# Generalized Additive Models
## Null model
### GAM 0: Null model
```{r, echo=FALSE, warning=FALSE}
#### 0. GAM: Null model ####
gam_null <- gam(richness ~ 1 + s(effort, bs = "re"), 
                data = cdata, 
                family = nb(link = "log"))
summary(gam_null)

## 2. AIC
aic_gam_null <- AIC(gam_null)
```

## GAM model
### GAM: rich ~ SVI + prop_arb
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SVI + prop_arb ####
## 1. Model
gam_SVI_prop_arb <- gam(richness ~ s(SVI) + s(prop_arb) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SVI_prop_arb)

## 2. AIC
aic_gam_SVI_prop_arb <- AIC(gam_SVI_prop_arb)
```

```{r, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SVI and prop_arb
sim_data3 <- expand.grid(SVI = seq(0, 1, length.out = 200),
                         prop_arb = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SVI_prop_arb
sim_data3 <- sim_data3 %>% 
  mutate(predicted_richness = predict(gam_SVI_prop_arb, newdata = transform(sim_data3, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_3 <- ggplot(data = sim_data3, aes(x = SVI, y = prop_arb, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SVI", y = "prop_arb") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of additive model")

plot_heatmap_3
```

### GAM: rich ~ SVI * prop_arb
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SVI * prop_arb ####
## 1. Model
gam_SVI_prop_arb_int <- gam(richness ~ s(SVI,prop_arb) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SVI_prop_arb_int)

## 2. AIC
aic_gam_SVI_prop_arb_int <- AIC(gam_SVI_prop_arb_int)
```

```{r, echo=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SVI and prop_arb
sim_data8 <- expand.grid(SVI = seq(0, 1, length.out = 200),
                         prop_arb = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SVI_prop_arb
sim_data8 <- sim_data8 %>% 
  mutate(predicted_richness = predict(gam_SVI_prop_arb_int, newdata = transform(sim_data8, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_8 <- ggplot(data = sim_data8, aes(x = SVI, y = prop_arb, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SVI", y = "prop_arb") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of interaction model")

plot_heatmap_8
```

# Comparing models
## AIC selection
```{r, echo=FALSE, warning=FALSE}
# Store model names and AIC values in a data frame
aic_values <- data.frame(
  Model = c("GLM Model (SVI + prop_arb)",
            "GLM Model (SVI * prop_arb)", 
            "GAM Model (SVI + prop_arb)", 
            "GAM Model (SVI * prop_arb)"),
  AIC = c(aic_glm_SVI_prop_arb,
          aic_glm_SVI_prop_arb_int,
          aic_gam_SVI_prop_arb, 
          aic_gam_SVI_prop_arb_int)
)

# Calculate delta AIC
aic_min <- min(aic_values$AIC)
aic_values$Delta_AIC <- aic_values$AIC - aic_min

# Sort by AIC
aic_values_sorted <- aic_values[order(aic_values$AIC), ]

# Print the sorted AIC values along with Delta AIC
print(aic_values_sorted)
```

# GLM check
Close to normality of the residuals and no heteroscedasticity pattern (non-constant variation of residuals).

```{r, echo=FALSE, warning=FALSE}
# Função para criar o gráfico de resíduos com linha reta em y = 0
create_residual_plot <- function(model, title) {
  residuals_data <- data.frame(
    Fitted = fitted.values(model),
    Residuals = residuals(model)
  )
  
  ggplot(residuals_data, aes(x = Fitted, y = Residuals)) +
    geom_point(alpha = 0.7, color = "black") +
    geom_hline(yintercept = 0, color = "red") +  # Linha reta em y = 0
    labs(x = "Fitted Values", y = "Residuals", title = title) +
    theme_classic() +  
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"), 
      panel.border = element_blank(),  
      axis.line = element_line(color = "black") 
    )
}

# Criando os 4 gráficos de resíduos com títulos ajustados
p1 <- create_residual_plot(glm_SVI_prop_arb, "GLM SVI + prop_arb")
p2 <- create_residual_plot(glm_SVI_prop_arb_int, "GLM SVI * prop_arb")

# Organizando os gráficos em um layout 2x2 usando patchwork
layout <- (p1 | p2)
layout

# Função para criar o Q-Q plot com linha de qqline do R base replicada
create_qq_plot <- function(model, title) {
  residuals_data <- data.frame(
    Residuals = sort(residuals(model)),
    Theoretical = sort(qnorm(ppoints(length(residuals(model)))))
  )
  
  # Calcula os quantis para a linha de qqline
  slope <- diff(quantile(residuals_data$Residuals, c(0.25, 0.75))) / 
           diff(quantile(residuals_data$Theoretical, c(0.25, 0.75)))
  intercept <- quantile(residuals_data$Residuals, 0.25) - slope * quantile(residuals_data$Theoretical, 0.25)
  
  ggplot(residuals_data, aes(sample = Residuals)) +
    geom_point(aes(x = Theoretical, y = Residuals), color = "black", alpha = 0.7) +
    geom_abline(intercept = intercept, slope = slope, color = "red") +  # Linha ajustada ao estilo qqline
    labs(x = "Theoretical Quantiles", y = "Sample Quantiles", title = title) +
    theme_classic() +
    theme(
      axis.title = element_text(size = 14), 
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"),  
      panel.border = element_blank(),  
      axis.line = element_line(color = "black")  
    )
}

# Criando os 4 Q-Q plots com títulos ajustados e linha no estilo do R base
qq1 <- create_qq_plot(glm_SVI_prop_arb, "Q-Q Plot: GLM SVI + prop_arb")
qq2 <- create_qq_plot(glm_SVI_prop_arb_int, "Q-Q Plot: GLM SVI * prop_arb")

# Organizando os Q-Q plots em um layout 2x2 usando patchwork
layout_qq <- (qq1 | qq2)
layout_qq
```

```{r, echo=FALSE, warning=FALSE}
create_residual_histogram <- function(model, title) {
  residuals_data <- data.frame(Residuals = residuals(model))
  
  ggplot(residuals_data, aes(x = Residuals)) +
    geom_histogram(color = "black", fill = "gray", alpha = 0.6, bins = 30) +  # Histograma com 30 bins
    labs(x = "Residuals", y = "Frequency", title = title) +
    theme_classic() +  
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 12), 
      plot.title = element_text(size = 16, face = "bold"), 
      panel.border = element_blank(),  
      axis.line = element_line(color = "black") 
    )
}


# Criando os 4 histogramas dos resíduos
h1 <- create_residual_histogram(glm_SVI_prop_arb, "Histogram: GLM SVI + prop_arb")
h2 <- create_residual_histogram(glm_SVI_prop_arb_int, "Histogram: GLM SVI * prop_arb")


# Organizando os gráficos em um layout 2x2 usando patchwork
layout_histograms <- (h1 | h2)
layout_histograms
```

```{r, echo=FALSE, warning=FALSE}
# Teste de normalidade de Shapiro-Wilk
shapiro_test_SVI_prop_arb <- shapiro.test(residuals(glm_SVI_prop_arb))
shapiro_test_SVI_prop_arb_int <- shapiro.test(residuals(glm_SVI_prop_arb_int))

# Exibir resultados do teste
print(shapiro_test_SVI_prop_arb)
print(shapiro_test_SVI_prop_arb_int)
```

```{r, echo=FALSE, warning=FALSE}
# Simular resíduos
sim_resid_SVI_prop_arb <- simulateResiduals(fittedModel = glm_SVI_prop_arb)
sim_resid_SVI_prop_arb_int <- simulateResiduals(fittedModel = glm_SVI_prop_arb_int)

# Plotar resíduos simulados para diagnóstico visual
plot(sim_resid_SVI_prop_arb)
plot(sim_resid_SVI_prop_arb_int)
```

# GAM check
## GAM: rich ~ SVI + prop_arb
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SVI + prop_arb ####
## 1. Model
gam_SVI_prop_arb <- gam(richness ~ s(SVI) + s(prop_arb) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SVI_prop_arb)

## 2. AIC
aic_gam_SVI_prop_arb <- AIC(gam_SVI_prop_arb)

## 3. Plot
plot(gam_SVI_prop_arb, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_3 <- vis.gam(gam_SVI_prop_arb, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SVI_prop_arb)
gam.check(gam_SVI_prop_arb)
```

## GAM: rich ~ SVI * prop_arb
```{r, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SVI * prop_arb ####
## 1. Model
gam_SVI_prop_arb_int <- gam(richness ~ s(SVI,prop_arb) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))
summary(gam_SVI_prop_arb_int)

## 2. AIC
aic_gam_SVI_prop_arb_int <- AIC(gam_SVI_prop_arb_int)

## 3. Plot
plot(gam_SVI_prop_arb_int, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_8 <- vis.gam(gam_SVI_prop_arb_int, theta = 150, n.grid = 50, lwd = 0.4)

## 4. Check
k.check(gam_SVI_prop_arb_int)
gam.check(gam_SVI_prop_arb_int)


```

## GAM ANOVA
```{r, echo=FALSE, warning=FALSE}
anova(gam_SVI_prop_arb, gam_SVI_prop_arb_int, test = "F")
```

# Visualize
```{r, echo=FALSE, warning=FALSE}
ggplot(cdata, aes(x = prop_arb, y = richness, color = SVI)) +
  geom_point(alpha = 0.7, size = 2) + 
  geom_smooth(method = "gam", formula = y ~ s(x), color = "black", fill = "black", alpha = 0.05, size = 0.5) +
  labs(x = "Proportion of arboreal vegetation", y = "Richness", title = "Richness with sharing and proportion of arboreal vegetation") +
  theme_classic() +  
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12), 
    plot.title = element_text(size = 16, face = "bold"), 
    panel.border = element_blank(),  
    axis.line = element_line(color = "black") 
  )
```
