---
title: "Assessing the Impact of Urban Infrastructure on Bird Biodiversity"
author: "Bianca Neves, Irina Lerner, Douglas Cirino"
date: "2024-09-25"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: paper
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)       # For data visualization
library(dplyr)         # For data manipulation
library(PerformanceAnalytics) # For correlation matrix
library(tidyr)         # For data tidying
library(sf)            # For handling spatial data
library(sp)            # For handling spatial data
library(spdep)         # For spatial dependence analysis
library(ggeffects)    # For extracting and plotting model effects
library(mgcv)          # For fitting GAMs
library(viridis)       # For color scales
library(patchwork)    # combine plots
library(car)
library(lme4)
library(MASS)

custom_theme <- theme_minimal() +
  theme(text = element_text(size = 12),
        plot.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

```

# Data 4km²
This is a $4km^2$ grid over the urban area of the city of São Paulo
```{r data, include=FALSE, warning=FALSE}
shp <- st_read("Z:/Bianca/eBird/Data/2km_urb_f2.5km/richness_effort_v20_total_2km_urb_f2.5km_30min_10spp_sem_exot_equiv_groups.shp")
#shp <- st_read("C:/Users/Bianca/Documents/USP/Projeto/eBird/Data/1km/richness_effort_v16_total_1km_urb_f1km_30min_10spp.shp")
cdata <- shp


## 2. Rename columns
cdata <- cdata %>% 
  rename(
    richness = richness_s,
    effort = effort_s,
    Bheight_m = Bhght_m,
    Bheight_s = Bhght_s,
    Vheight_m = Vhght_m,
    Vheight_s = Vhght_s,
    Wdistance = Wdistnc,
    contig_f = contg_f,
    contig_h = contg_h,
    pland_f = plant_f,
    pland_h = pland_h,
    area_mn_f = ar_mn_f,
    area_mn_h = ar_mn_h,
    prop_veg = prop_vg,
    prop_for = prp_frs,
    prop_wet = prp__wt,
    prop_grc = prp_grc,
    prop_hrb = prp_hrb,
    landscape_ = lndscp_,
    contig_m_tot = cntg_m_,
    pland_tot = plnd_tt,
    nlsi_tot = nlsi_tt,
    area_m_tot = ar_mn_t,
    Gdistance = Gdistnc,
    geometry = geometry
  )

cdata <- cdata %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  arrange(id)

cdata$centx <- st_coordinates(st_centroid(cdata))[, 1]
cdata$centy <- st_coordinates(st_centroid(cdata))[, 2]

```

## Spatial correlation
```{r spatial, echo=FALSE, warning=FALSE}
# 1. Testing for spatial correlation
coords <- st_centroid(cdata)
nb <- poly2nb(cdata)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
moran.mc(cdata$richness, lw, nsim = 999)
# We see that there is a small spatial correlation

# 2. Check the spatial distribuition of richness
(ggplot(data = cdata) +
  geom_sf(aes(fill = richness), color = NA) +
  scale_fill_viridis_c(direction = -1) +
  custom_theme)
```
While the Moran's I value of 0.18297 may not seem exceptionally high at first glance, it is considered significant in spatial autocorrelation studies, indicating a clear and notable pattern of clustering in bird richness. 


#Variables
## Correlation between variables
```{r cor, echo=FALSE, warning=FALSE}
## 1. Predictors
predictors <- c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")

## 2. Correlation of variables
cor_mat_structure <- as.data.frame(cdata)[, c("richness", predictors)]
suppressWarnings(chart.Correlation(cor_mat_structure, histogram = TRUE, pch = 16, method = "pearson"))
```

## Multicollinearity in the predictors
To check for multicollinearity in the predictors used in the GAM, I calculated the Variance Inflation Factor (VIF) using a linear model (lm). Although the GAM is a more flexible model, the VIF cannot be directly applied to GAMs. Instead, I fit a linear regression model with the same predictors (SVI, prop_arb, and effort) as in the GAM. The VIF was then calculated from this linear model to assess the degree of multicollinearity between the variables.
```{r vif, echo=FALSE, warning=FALSE}
vif_model <- lm(richness ~ SVI + prop_arb + effort, data = cdata)
vif(vif_model)
```
The VIF results confirm that the predictor variables do not exhibit problematic multicollinearity, ensuring the robustness of the model estimates.

## Association between richness and variables
```{r rich, echo=FALSE, warning=FALSE}
## Association between richness and variables
# Reshape the data to long format with only 'value' and 'variable'
cdata_long <- cdata[, c("richness", "effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb")]
cdata_long <- cdata_long %>% 
  pivot_longer(
    cols = c("effort", "SVI", "SEVIo", "SUVI", "SGVI", "prop_arb"),
    names_to = "variable",
    values_to = "value"
  )

# Create a faceted plot for the specified predictors
(plot_structure <- ggplot(cdata_long, aes(x = value, y = richness)) + 
    geom_point(alpha = 0.7, color = "black") + 
    geom_smooth(method = "loess", color = "blue", fill = "blue", alpha = 0.1) +
    facet_wrap(~ variable, scales = "free_x", ncol = 2, nrow = 3) +  # Adjusting to 3 rows and 2 columns
    labs(x = "Value", y = "Richness") +
    theme_classic() + 
    theme(
      axis.title = element_text(size = 14),  
      axis.text = element_text(size = 14),
      strip.text = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 16, face = "bold")
    ) +
    ggtitle("Richness  vs. Predictors"))

```

# Generalized linear models
## Null model
### GLM 0: Null model
```{r, echo=FALSE, warning=FALSE}
# Fit the mixed model with effort as a random intercept
glm_null <- glmer.nb(richness ~ 1 + (1 | effort), 
                     data = cdata)

summary(glm_null)

summary_glm_null <- summary(glm_null)
estimate_glm_null <- summary_glm_null$coefficients[, "Estimate"]
se_glm_null <- summary_glm_null$coefficients[, "Std. Error"]
pval_glm_null <- summary_glm_null$coefficients[, "Pr(>|z|)"]
df_glm_null <- summary_glm_null$df.residual

aic_glm_null <- AIC(glm_null)
print(aic_glm_null)
```

## GLM with one predictor
### GLM: rich ~ SVI
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with SVI as a fixed effect and effort as a random effect
glm_SVI <- glmer.nb(richness ~ SVI + (1 | effort), data = cdata)

summary(glm_SVI)

summary_glm_SVI <- summary(glm_SVI)
estimate_glm_SVI <- summary_glm_SVI$coefficients[, "Estimate"]
se_glm_SVI <- summary_glm_SVI$coefficients[, "Std. Error"]
pval_glm_SVI <- summary_glm_SVI$coefficients[, "Pr(>|z|)"]
df_glm_SVI <- summary_glm_SVI$df.residual

aic_glm_SVI <- AIC(glm_SVI)
print(aic_glm_SVI)
```

### GAM 5: rich ~ prop_arb
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with prop_arb as a fixed effect and effort as a random effect
glm_prop_arb <- glmer.nb(richness ~ prop_arb + (1 | effort), data = cdata)

summary(glm_prop_arb)

summary_glm_prop_arb <- summary(glm_prop_arb)
estimate_glm_prop_arb <- summary_glm_prop_arb$coefficients[, "Estimate"]
se_glm_prop_arb <- summary_glm_prop_arb$coefficients[, "Std. Error"]
pval_glm_prop_arb <- summary_glm_prop_arb$coefficients[, "Pr(>|z|)"]
df_glm_prop_arb <- summary_glm_prop_arb$df.residual

aic_glm_prop_arb <- AIC(glm_prop_arb)
print(aic_glm_prop_arb)
```

## GAM with additive effects
### GLM: rich ~ SVI + prop_arb
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with SVI and prop_arb as fixed effects and effort as a random effect
glm_svi_prop_arb <- glmer.nb(richness ~ SVI + prop_arb + (1 | effort), data = cdata)

summary(glm_svi_prop_arb)

summary_glm_svi_prop_arb <- summary(glm_svi_prop_arb)
estimate_glm_svi_prop_arb <- summary_glm_svi_prop_arb$coefficients[, "Estimate"]
se_glm_svi_prop_arb <- summary_glm_svi_prop_arb$coefficients[, "Std. Error"]
pval_glm_svi_prop_arb <- summary_glm_svi_prop_arb$coefficients[, "Pr(>|z|)"]
df_glm_svi_prop_arb <- summary_glm_svi_prop_arb$df.residual

aic_glm_svi_prop_arb <- AIC(glm_svi_prop_arb)
print(aic_glm_svi_prop_arb)
```

### GLM: rich ~ SEVIo + SGVI 
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with SEVIo and SGVI as fixed effects and effort as a random effect
glm_sevio_sgvi <- glmer.nb(richness ~ SEVIo + SGVI + (1 | effort), data = cdata)

summary(glm_sevio_sgvi)

summary_glm_sevio_sgvi <- summary(glm_sevio_sgvi)
estimate_glm_sevio_sgvi <- summary_glm_sevio_sgvi$coefficients[, "Estimate"]
se_glm_sevio_sgvi <- summary_glm_sevio_sgvi$coefficients[, "Std. Error"]
pval_glm_sevio_sgvi <- summary_glm_sevio_sgvi$coefficients[, "Pr(>|z|)"]
df_glm_sevio_sgvi <- summary_glm_sevio_sgvi$df.residual

aic_glm_sevio_sgvi <- AIC(glm_sevio_sgvi)
print(aic_glm_sevio_sgvi)
```





### GAM 3: rich ~ SEVIo + SUVI
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with SEVIo and SUVI as fixed effects and effort as a random effect
glm_sevio_suvi <- glmer.nb(richness ~ SEVIo + SUVI + (1 | effort), data = cdata)

summary(glm_sevio_suvi)

summary_glm_sevio_suvi <- summary(glm_sevio_suvi)
estimate_glm_sevio_suvi <- summary_glm_sevio_suvi$coefficients[, "Estimate"]
se_glm_sevio_suvi <- summary_glm_sevio_suvi$coefficients[, "Std. Error"]
pval_glm_sevio_suvi <- summary_glm_sevio_suvi$coefficients[, "Pr(>|z|)"]
df_glm_sevio_suvi <- summary_glm_sevio_suvi$df.residual

aic_glm_sevio_suvi <- AIC(glm_sevio_suvi)
print(aic_glm_sevio_suvi)
```




## GLM with interaction effects
### GLM: rich ~ SVI * prop_arb
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with interaction between SVI and prop_arb as fixed effects and effort as a random effect
glm_svi_prop_arb_int <- glmer.nb(richness ~ SVI * prop_arb + (1 | effort), data = cdata)

summary(glm_svi_prop_arb_int)

summary_glm_svi_prop_arb_int <- summary(glm_svi_prop_arb_int)
estimate_glm_svi_prop_arb_int <- summary_glm_svi_prop_arb_int$coefficients[, "Estimate"]
se_glm_svi_prop_arb_int <- summary_glm_svi_prop_arb_int$coefficients[, "Std. Error"]
pval_glm_svi_prop_arb_int <- summary_glm_svi_prop_arb_int$coefficients[, "Pr(>|z|)"]
df_glm_svi_prop_arb_int <- summary_glm_svi_prop_arb_int$df.residual

aic_glm_svi_prop_arb_int <- AIC(glm_svi_prop_arb_int)
print(aic_glm_svi_prop_arb_int)
```


### GLM: rich ~ SEVIo * SGVI 
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with interaction between SEVIo and SGVI as fixed effects and effort as a random effect
glm_sevio_sgvi_int <- glmer.nb(richness ~ SEVIo * SGVI + (1 | effort), data = cdata)

summary(glm_sevio_sgvi_int)

summary_glm_sevio_sgvi_int <- summary(glm_sevio_sgvi_int)
estimate_glm_sevio_sgvi_int <- summary_glm_sevio_sgvi_int$coefficients[, "Estimate"]
se_glm_sevio_sgvi_int <- summary_glm_sevio_sgvi_int$coefficients[, "Std. Error"]
pval_glm_sevio_sgvi_int <- summary_glm_sevio_sgvi_int$coefficients[, "Pr(>|z|)"]
df_glm_sevio_sgvi_int <- summary_glm_sevio_sgvi_int$df.residual

aic_glm_sevio_sgvi_int <- AIC(glm_sevio_sgvi_int)
print(aic_glm_sevio_sgvi_int)
```

### GLM: rich ~ SEVIo * SUVI
```{r, echo=FALSE, warning=FALSE}
# Fit the GLM with interaction between SEVIo and SUVI as fixed effects and effort as a random effect
glm_sevio_suvi_int <- glmer.nb(richness ~ SEVIo * SUVI + (1 | effort), data = cdata)

summary(glm_sevio_suvi_int)

summary_glm_sevio_suvi_int <- summary(glm_sevio_suvi_int)
estimate_glm_sevio_suvi_int <- summary_glm_sevio_suvi_int$coefficients[, "Estimate"]
se_glm_sevio_suvi_int <- summary_glm_sevio_suvi_int$coefficients[, "Std. Error"]
pval_glm_sevio_suvi_int <- summary_glm_sevio_suvi_int$coefficients[, "Pr(>|z|)"]
df_glm_sevio_suvi_int <- summary_glm_sevio_suvi_int$df.residual

aic_glm_sevio_suvi_int <- AIC(glm_sevio_suvi_int)
print(aic_glm_sevio_suvi_int)
```

## Comparing GLM models
```{r, echo=FALSE, warning=FALSE}
# Store model names and AIC values in a data frame
aic_values_glm <- data.frame(
  Model = c("Model 1 (SEVIo + SUVI)", 
            "Model 2 (SEVIo * SUVI)", 
            "Model 3 (SEVIo + SGVI)",  
            "Model 4 (SEVIo * SGVI)",  
            "Model 5 (SVI + prop_arb)", 
            "Model 6 (SVI * prop_arb)", 
            "Model 7 (prop_arb)",
            "Null Model"),
  AIC = c(aic_glm_sevio_suvi, 
          aic_glm_sevio_suvi_int, 
          aic_glm_sevio_sgvi,     
          aic_glm_sevio_sgvi_int,  
          aic_glm_svi_prop_arb, 
          aic_glm_svi_prop_arb_int, 
          aic_glm_prop_arb,
          aic_glm_null)
)

# Calculate delta AIC
aic_min_glm <- min(aic_values_glm$AIC)
aic_values_glm$Delta_AIC_glm <- aic_values_glm$AIC - aic_min_glm  # Use aic_min_glm here

# Sort by AIC
aic_values_glm_sorted <- aic_values_glm[order(aic_values_glm$AIC), ]

# Print the sorted AIC values along with Delta AIC
print(aic_values_glm_sorted)

```



# Generalized additive models
## Null model
### GAM 0: Null model
```{r, echo=FALSE, warning=FALSE}
#### 0. GAM: Null model ####
gam_null <- gam(richness ~ 1 + s(effort, bs = "re"), 
                data = cdata, 
                family = nb(link = "log"))

summary(gam_null)

summary_gam_null <- summary(gam_null)
estimate_gam_null <- summary_gam_null$p.table[, "Estimate"]
se_gam_null <- summary_gam_null$p.table[, "Std. Error"]
pval_gam_null <- summary_gam_null$p.table[, "Pr(>|z|)"]
df_gam_null <- summary_gam_null$df.residual

## 2. AIC
aic_gam_null <- AIC(gam_null)
AIC(gam_null)
```

## GAM with one predictor
### GAM 4: rich ~ SVI 
```{r, echo=FALSE, warning=FALSE}
## 1. Model
gam_SVI <- gam(richness ~ s(SVI) + s(effort, bs = "re"), 
               data = cdata, 
               family = nb(link = "log"))

summary(gam_SVI)

summary_SVI <- summary(gam_SVI)
estimate_SVI <- summary_SVI$p.table[, "Estimate"]
se_SVI <- summary_SVI$p.table[, "Std. Error"]
pval_SVI <- summary_SVI$p.table[, "Pr(>|z|)"]
df_SVI <- summary(gam_SVI)$df.residual

## 2. AIC
aic_SVI <- AIC(gam_SVI)
AIC(gam_SVI)

## 3. Plot
plot(gam_SVI, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_4 <- vis.gam(gam_SVI, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SVI
sim_data4 <- expand.grid(SVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SVI
sim_data4 <- sim_data4 %>% 
  mutate(predicted_richness = predict(gam_SVI, newdata = transform(sim_data4, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_4 <- ggplot(data = sim_data4, aes(x = SVI, y = predicted_richness, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SVI", y = "Predicted richness") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 4")
```

### GAM 5: rich ~ prop_arb
```{r, echo=FALSE, warning=FALSE}
## 1. Model
gam_prop_arb <- gam(richness ~ s(prop_arb) + s(effort, bs = "re"), 
                    data = cdata, 
                    family = nb(link = "log"))

summary_prop_arb <- summary(gam_prop_arb)
estimate_prop_arb <- summary_prop_arb$p.table[, "Estimate"]
se_prop_arb <- summary_prop_arb$p.table[, "Std. Error"]
pval_prop_arb <- summary_prop_arb$p.table[, "Pr(>|z|)"]
df_prop_arb <- summary(gam_prop_arb)$df.residual

## 2. AIC
aic_prop_arb <- AIC(gam_prop_arb)

## 3. Plot
plot(gam_prop_arb, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_5 <- vis.gam(gam_prop_arb, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r gam5p, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for prop_arb
sim_data5 <- expand.grid(prop_arb = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_prop_arb
sim_data5 <- sim_data5 %>% 
  mutate(predicted_richness = predict(gam_prop_arb, newdata = transform(sim_data5, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_5 <- ggplot(data = sim_data5, aes(x = prop_arb, y = predicted_richness, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "Proportion of vegetation", y = "Predicted richness") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 5")
```

## GAM with additive effects
### GAM 1: rich ~ SVI + prop_arb
```{r gam1, echo=FALSE, warning=FALSE}
#### 1. GAM: rich ~ SVI + prop_arb ####
## 1. Model
gam_SVI_prop_arb <- gam(richness ~ s(SVI) + s(prop_arb) + s(effort, bs = "re"), 
                    data = cdata, 
                    family = nb(link = "log"))

summary_SVI_prop_arb <- summary(gam_SVI_prop_arb)
estimate_SVI_prop_arb <- summary_SVI_prop_arb$p.table[, "Estimate"]
se_SVI_prop_arb <- summary_SVI_prop_arb$p.table[, "Std. Error"]
pval_SVI_prop_arb <- summary_SVI_prop_arb$p.table[, "Pr(>|z|)"]
df_SVI_prop_arb <- summary(gam_SVI_prop_arb)$df.residual

## 2. AIC
aic_SVI_prop_arb <- AIC(gam_SVI_prop_arb)

## 3. Plot
plot(gam_SVI_prop_arb, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_1 <- vis.gam(gam_SVI_prop_arb, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r gam1p, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIn and SVI
sim_data1 <- expand.grid(prop_arb = seq(0, 1, length.out = 200),
                        SVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)  # Example: using the median of `effort`

# 4.2: Use predict() to generate predictions from the model gam_SVI_prop_arb
sim_data1 <- sim_data1 %>% 
  mutate(predicted_richness = predict(gam_SVI_prop_arb, newdata = transform(sim_data1, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_1 <- ggplot(data = sim_data1, aes(x = SVI, y = prop_arb, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SVI", y = "Proportion of vegetation") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 1")

print(plot_heatmap_1)
```

### GAM 2: rich ~ SEVIo + SGVI 
```{r gam2, echo=FALSE, warning=FALSE}
#### 2. GAM: rich ~ SEVIo + SGVI ####
## 1. Model
gam_SEVIo_SGVI <- gam(richness ~ s(SEVIo) + s(SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))

summary_SEVIo_SGVI <- summary(gam_SEVIo_SGVI)
estimate_SEVIo_SGVI <- summary_SEVIo_SGVI$p.table[, "Estimate"]
se_SEVIo_SGVI <- summary_SEVIo_SGVI$p.table[, "Std. Error"]
pval_SEVIo_SGVI <- summary_SEVIo_SGVI$p.table[, "Pr(>|z|)"]
df_SEVIo_SGVI <- summary(gam_SEVIo_SGVI)$df.residual

## 2. AIC
aic_SEVIo_SGVI <- AIC(gam_SEVIo_SGVI)

## 3. Plot
plot(gam_SEVIo_SGVI, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_2 <- vis.gam(gam_SEVIo_SGVI, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r gam2p, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SGVI
sim_data2 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SGVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)  # Example: using the median of `effort`

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SGVI
sim_data2 <- sim_data2 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SGVI, newdata = transform(sim_data2, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_2 <- ggplot(data = sim_data2, aes(x = SEVIo, y = SGVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SGVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 2")
```

### GAM 3: rich ~ SEVIo + SUVI
```{r gam3, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo + SUVI ####
## 1. Model
gam_SEVIo_SUVI <- gam(richness ~ s(SEVIo) + s(SUVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))

summary_SEVIo_SUVI <- summary(gam_SEVIo_SUVI)
estimate_SEVIo_SUVI <- summary_SEVIo_SUVI$p.table[, "Estimate"]
se_SEVIo_SUVI <- summary_SEVIo_SUVI$p.table[, "Std. Error"]
pval_SEVIo_SUVI <- summary_SEVIo_SUVI$p.table[, "Pr(>|z|)"]
df_SEVIo_SUVI <- summary(gam_SEVIo_SUVI)$df.residual

## 2. AIC
aic_SEVIo_SUVI <- AIC(gam_SEVIo_SUVI)

## 3. Plot
plot(gam_SEVIo_SUVI, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_3 <- vis.gam(gam_SEVIo_SUVI, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r gam3p, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SUVI
sim_data3 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SUVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SUVI
sim_data3 <- sim_data3 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SUVI, newdata = transform(sim_data3, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_3 <- ggplot(data = sim_data3, aes(x = SEVIo, y = SUVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SUVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 3")
```

## GAM with interaction effects
### GAM 6: rich ~ SVI * prop_arb
```{r gam6, echo=FALSE, warning=FALSE}
#### 1. GAM: rich ~ SVI * prop_arb ####
## 1. Model
gam_SVI_prop_arb_int <- gam(richness ~ s(SVI,prop_arb) + s(effort, bs = "re"), 
                    data = cdata, 
                    family = nb(link = "log"))

summary_SVI_prop_arb_int <- summary(gam_SVI_prop_arb_int)
estimate_SVI_prop_arb_int <- summary_SVI_prop_arb_int$p.table[, "Estimate"]
se_SVI_prop_arb_int <- summary_SVI_prop_arb_int$p.table[, "Std. Error"]
pval_SVI_prop_arb_int <- summary_SVI_prop_arb_int$p.table[, "Pr(>|z|)"]
df_SVI_prop_arb_int <- summary(gam_SVI_prop_arb_int)$df.residual

## 2. AIC
aic_SVI_prop_arb_int <- AIC(gam_SVI_prop_arb_int)

## 3. Plot
plot(gam_SVI_prop_arb_int, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_6 <- vis.gam(gam_SVI_prop_arb_int, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r gam6p, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for prop_arb and SVI
sim_data6 <- expand.grid(prop_arb = seq(0, 1, length.out = 200),
                        SVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)  # Example: using the median of `effort`

# 4.2: Use predict() to generate predictions from the model gam_SVI_prop_arb
sim_data6 <- sim_data6 %>% 
  mutate(predicted_richness = predict(gam_SVI_prop_arb_int, newdata = transform(sim_data6, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_6 <- ggplot(data = sim_data6, aes(x = SVI, y = prop_arb, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SVI", y = "Proportion of vegetation") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 6")

print(plot_heatmap_6)
```

### GAM 7: rich ~ SEVIo * SGVI 
```{r gam7, echo=FALSE, warning=FALSE}
#### 7. GAM: rich ~ SEVIo * SGVI ####
## 1. Model
gam_SEVIo_SGVI_int <- gam(richness ~ s(SEVIo,SGVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))

summary_SEVIo_SGVI_int <- summary(gam_SEVIo_SGVI_int)
estimate_SEVIo_SGVI_int <- summary_SEVIo_SGVI_int$p.table[, "Estimate"]
se_SEVIo_SGVI_int <- summary_SEVIo_SGVI_int$p.table[, "Std. Error"]
pval_SEVIo_SGVI_int <- summary_SEVIo_SGVI_int$p.table[, "Pr(>|z|)"]
df_SEVIo_SGVI_int <- summary(gam_SEVIo_SGVI_int)$df.residual

## 2. AIC
aic_SEVIo_SGVI_int <- AIC(gam_SEVIo_SGVI_int)

## 3. Plot
plot(gam_SEVIo_SGVI_int, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_7 <- vis.gam(gam_SEVIo_SGVI_int, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r gam7p, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SGVI
sim_data7 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SGVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)  # Example: using the median of `effort`

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SGVI
sim_data7 <- sim_data7 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SGVI_int, newdata = transform(sim_data7, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_7 <- ggplot(data = sim_data7, aes(x = SEVIo, y = SGVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SGVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 7")
```

### GAM 8: rich ~ SEVIo * SUVI
```{r gam8, echo=FALSE, warning=FALSE}
#### 3. GAM: rich ~ SEVIo * SUVI ####
## 1. Model
gam_SEVIo_SUVI_int <- gam(richness ~ s(SEVIo,SUVI) + s(effort, bs = "re"), 
                      data = cdata, 
                      family = nb(link = "log"))

summary_SEVIo_SUVI_int <- summary(gam_SEVIo_SUVI_int)
estimate_SEVIo_SUVI_int <- summary_SEVIo_SUVI_int$p.table[, "Estimate"]
se_SEVIo_SUVI_int <- summary_SEVIo_SUVI_int$p.table[, "Std. Error"]
pval_SEVIo_SUVI_int <- summary_SEVIo_SUVI_int$p.table[, "Pr(>|z|)"]
df_SEVIo_SUVI_int <- summary(gam_SEVIo_SUVI_int)$df.residual

## 2. AIC
aic_SEVIo_SUVI_int <- AIC(gam_SEVIo_SUVI_int)

## 3. Plot
plot(gam_SEVIo_SUVI_int, pages = 1, residuals = TRUE, rug = TRUE)
plot_vis_gam_8 <- vis.gam(gam_SEVIo_SUVI_int, theta = 150, n.grid = 50, lwd = 0.4)
```

```{r gam8p, include=FALSE, warning=FALSE}
## 4. Heatmap
# 4.1 Create a grid of values for SEVIo and SUVI
sim_data8 <- expand.grid(SEVIo = seq(0, 1, length.out = 200),
                         SUVI = seq(0, 1, length.out = 200))

# Add a fixed value for `effort` to the grid
fixed_effort <- median(cdata$effort)

# 4.2: Use predict() to generate predictions from the model gam_SEVIo_SUVI
sim_data8 <- sim_data8 %>% 
  mutate(predicted_richness = predict(gam_SEVIo_SUVI_int, newdata = transform(sim_data8, effort = fixed_effort), type = "response"))

# 4.3: Plot the results using geom_tile() for a heatmap
plot_heatmap_8 <- ggplot(data = sim_data8, aes(x = SEVIo, y = SUVI, fill = predicted_richness)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Predicted richness") +
  labs(x = "SEVIo", y = "SUVI") +
  theme_classic() + 
  theme(
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ggtitle("Heatmap of model 8")
```


## Comparing GAM models

```{r compar, echo=FALSE, warning=FALSE}
# Store model names and AIC values in a data frame
aic_values <- data.frame(
  Model = c("Model 1 (SVI + prop_arb)", 
            "Model 2 (SEVIo + SGVI)", 
            "Model 3 (SEVIo + SUVI)", 
            "Model 4 (SVI)", 
            "Model 5 (prop_arb)",
            "Model 6 (SVI * prop_arb)", 
            "Model 7 (SEVIo * SGVI)", 
            "Model 8 (SEVIo * SUVI)",
            "Null Model"),
  AIC = c(aic_SVI_prop_arb, 
          aic_SEVIo_SGVI, 
          aic_SEVIo_SUVI, 
          aic_SVI, 
          aic_prop_arb,
          aic_SVI_prop_arb_int, 
          aic_SEVIo_SGVI_int, 
          aic_SEVIo_SUVI_int,
          aic_gam_null)
)

# Calculate delta AIC
aic_min <- min(aic_values$AIC)
aic_values$Delta_AIC <- aic_values$AIC - aic_min

# Sort by AIC
aic_values_sorted <- aic_values[order(aic_values$AIC), ]

# Print the sorted AIC values along with Delta AIC
print(aic_values_sorted)
```

```{r heatmap, echo=FALSE, warning=FALSE}
plot_heatmap_1
plot_heatmap_6
plot_heatmap_3
plot_heatmap_2
plot_heatmap_7
plot_heatmap_8
plot_heatmap_5
plot_heatmap_4
```






# Model selection
```{r, echo=FALSE, warning=FALSE}
aic_glm_gam <- data.frame(
  Model = c("Model GLM (SEVIo + SUVI)", 
            "Model GLM (SEVIo * SUVI)", 
            "Model GLM (SEVIo + SGVI)",  
            "Model GLM (SEVIo * SGVI)",  
            "Model GLM (SVI + prop_arb)", 
            "Model GLM (SVI * prop_arb)", 
            "Model GLM (prop_arb)",
            "Model GAM (SVI + prop_arb)", 
            "Model GAM (SEVIo + SGVI)", 
            "Model GAM (SEVIo + SUVI)", 
            "Model GAM (SVI)", 
            "Model GAM (prop_arb)",
            "Model GAM (SVI * prop_arb)", 
            "Model GAM (SEVIo * SGVI)", 
            "Model GAM (SEVIo * SUVI)"),
  AIC = c(aic_glm_sevio_suvi, 
          aic_glm_sevio_suvi_int, 
          aic_glm_sevio_sgvi,     
          aic_glm_sevio_sgvi_int,  
          aic_glm_svi_prop_arb, 
          aic_glm_svi_prop_arb_int, 
          aic_glm_prop_arb,
          aic_SVI_prop_arb, 
          aic_SEVIo_SGVI, 
          aic_SEVIo_SUVI, 
          aic_SVI, 
          aic_prop_arb,
          aic_SVI_prop_arb_int, 
          aic_SEVIo_SGVI_int, 
          aic_SEVIo_SUVI_int)
)

# Calculate delta AIC
aic_min_glm_gam <- min(aic_glm_gam$AIC)
aic_glm_gam$Delta_AIC_glm_gam <- aic_glm_gam$AIC - aic_min_glm_gam

# Sort by AIC
aic_glm_gam_sorted <- aic_glm_gam[order(aic_glm_gam$AIC), ]

# Print the sorted AIC values along with Delta AIC
print(aic_glm_gam_sorted)
```
