---
title: "Exploring eBird data from 2015 to 2023"
author: "Bianca Neves"
date: "24/09/24"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: paper
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(tidyr)
library(sf, sp)
library(patchwork)
library(mgcv)

```

# Data
The data is from eBird for the period from 2015 to 2023, covering the urban area of the city of São Paulo. Only cells with 30 minutes or more of observation and 10 or more bird species were selected.

```{r data, include=FALSE}
# 1km
shp_1 <- st_read("Z:/Bianca/eBird/Data/1km_urb_f1km/richness_effort_v20_total_1km_urb_f1km_30min_10spp_sem_exot_equiv_groups.shp")
data_1km <- shp_1
data_1km$scale <- "1 km²" 
data_1km <- data_1km %>% 
  rename(
    richness = richness_s,
    effort = effort_s,
    Bheight_m = Bhght_m,
    Bheight_s = Bhght_s,
    Vheight_m = Vhght_m,
    Vheight_s = Vhght_s,
    Wdistance = Wdistnc,
    contig_f = contg_f,
    contig_h = contg_h,
    pland_f = plant_f,
    pland_h = pland_h,
    area_mn_f = ar_mn_f,
    area_mn_h = ar_mn_h,
    prop_veg = prop_vg,
    prop_for = prp_frs,
    prop_wet = prp_wtl,
    prop_grc = prp_grc,
    prop_hrb = prp_hrb,
    landscape_ = lndscp_,
    contig_m_tot = cntg_m_,
    pland_tot = plnd_tt,
    nlsi_tot = nlsi_tt,
    area_m_tot = ar_mn_t,
    Gdistance = Gdistnc,
    geometry = geometry
  )

data_1km <- data_1km %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  arrange(id)

#4km
shp_4 <- st_read("Z:/Bianca/eBird/Data/2km_urb_f2.5km/richness_effort_v20_total_2km_urb_f2.5km_30min_10spp_sem_exot_equiv_groups.shp")
data_4km <- shp_4
data_4km$scale <- "4 km²"

data_4km <- data_4km %>% 
  rename(
    richness = richness_s,
    effort = effort_s,
    Bheight_m = Bhght_m,
    Bheight_s = Bhght_s,
    Vheight_m = Vhght_m,
    Vheight_s = Vhght_s,
    Wdistance = Wdistnc,
    contig_f = contg_f,
    contig_h = contg_h,
    pland_f = plant_f,
    pland_h = pland_h,
    area_mn_f = ar_mn_f,
    area_mn_h = ar_mn_h,
    prop_veg = prop_vg,
    prop_for = prp_frs,
    prop_wet = prp__wt,
    prop_grc = prp_grc,
    prop_hrb = prp_hrb,
    landscape_ = lndscp_,
    contig_m_tot = cntg_m_,
    pland_tot = plnd_tt,
    nlsi_tot = nlsi_tt,
    area_m_tot = ar_mn_t,
    Gdistance = Gdistnc,
    geometry = geometry
  )

data_4km <- data_4km %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  arrange(id)

# 9km
shp_9 <- st_read("Z:/Bianca/eBird/Data/3km_urb_f5km/richness_effort_v20_total_3km_urb_f5km_30min_10spp_sem_exot_equiv_groups.shp")
data_9km <- shp_9
data_9km$scale <- "9 km²"

data_9km <- data_9km %>% 
  rename(
    richness = richness_s,
    effort = effort_s,
    Bheight_m = Bhght_m,
    Bheight_s = Bhght_s,
    Vheight_m = Vhght_m,
    Vheight_s = Vhght_s,
    Wdistance = Wdistnc,
    contig_f = contg_f,
    contig_h = contg_h,
    pland_f = plant_f,
    pland_h = pland_h,
    area_mn_f = ar_mn_f,
    area_mn_h = ar_mn_h,
    prop_veg = prop_vg,
    prop_for = prp_frs,
    prop_wet = prp__wt,
    prop_grc = prp_grc,
    prop_hrb = prp_hrb,
    landscape_ = lndscp_,
    contig_m_tot = cntg_m_,
    pland_tot = plnd_tt,
    nlsi_tot = nlsi_tt,
    area_m_tot = ar_mn_t,
    Gdistance = Gdistnc,
    geometry = geometry
  )

data_9km <- data_9km %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  arrange(id)

# Combine datasets into one data frame
combined_data <- bind_rows(data_1km, data_4km, data_9km)
```

## Richness
### Density
```{r, echo=FALSE}
ggplot(combined_data, aes(x = richness, fill = scale)) +
  geom_density(alpha = 0.5) +
  labs(x = "Richness", y = "Density") +
  ggtitle("Species richness by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```
```{r, echo=FALSE}
ggplot(combined_data, aes(x = richness, fill = scale)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), position = "dodge", binwidth = 5, alpha = 0.7, color = "black") +
  labs(x = "Richness", y = "Proportion") +
  ggtitle("Species richness by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```
```{r, echo=FALSE}  
# Boxplot for species richness
ggplot(combined_data, aes(x = scale, y = richness, fill = scale)) +
  geom_boxplot() +
  geom_jitter(color = "black", size = 2, alpha = 0.1) +
  labs(x = "Scale", y = "Richness") +
  ggtitle("Species richness by scale") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```
```{r, echo=FALSE}
# Calcular médias e variâncias
mean_1km <- mean(data_1km$richness)
var_1km <- var(data_1km$richness)

mean_4km <- mean(data_4km$richness)
var_4km <- var(data_4km$richness)

mean_9km <- mean(data_9km$richness)
var_9km <- var(data_9km$richness)

# Criar tabela
table_richness <- data.frame(
  Scale = c("1km²", "4km²", "9km²"),
  Mean = c(mean_1km, mean_4km, mean_9km),
  Variance = c(var_1km, var_4km, var_9km)
)

# Visualizar tabela
print(table_richness)
```

# Time effort
Sum of the total time in minutes of independent sampling (filtered by person and by group).
```{r, echo=FALSE}
ggplot(combined_data, aes(x = effort * 60, fill = scale)) +
  geom_density(alpha = 0.5) +
  labs(x = "Effort (hours)", y = "Density") +
  ggtitle("Time effort by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```
```{r, echo=FALSE}
ggplot(combined_data, aes(x = effort * 60, fill = scale)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), position = "dodge", binwidth = 60, alpha = 0.7, color = "black") +
  labs(x = "Effort (hours)", y = "Proportion") +
  ggtitle("Time effort by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```

```{r, echo=FALSE}  
# Boxplot for effort
ggplot(combined_data, aes(x = scale, y = effort * 60, fill = scale)) +
  geom_boxplot() +
  geom_jitter(color = "black", size = 2, alpha = 0.1) +
  labs(x = "Scale", y = "Effort (hours)") +
  ggtitle("Time effort by scale") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```

```{r, echo=FALSE}
# Histogram with binwidth = 100 and y as proportion, colored by scale
ggplot(combined_data, aes(x = effort / 60, fill = scale)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 100, color = "black", alpha = 0.7) + 
  labs(x = "Effort (hour)", y = "Proportion") +
  theme_classic() + 
  theme(
    legend.position = "top",
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  ggtitle("Time effort by scale")

# Histogram with binwidth = 10 and xlim adjusted to 0-100 and y as proportion, colored by scale
ggplot(combined_data, aes(x = effort / 60, fill = scale)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 10, color = "black", alpha = 0.7) + 
  labs(x = "Effort (hour)", y = "Proportion") +
  coord_cartesian(xlim = c(0, 100)) +
  theme_classic() + 
  theme(
    legend.position = "top",
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  ggtitle("Time effort by scale")

# Histogram with binwidth = 1 and xlim adjusted to 0-50 and y as proportion, colored by scale
ggplot(combined_data, aes(x = effort / 60, fill = scale)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 1, color = "black", alpha = 0.7) + 
  labs(x = "Effort (hour)", y = "Proportion") +
  coord_cartesian(xlim = c(0, 50)) +
  theme_classic() + 
  theme(
    legend.position = "top",
    axis.title = element_text(size = 14),  
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  ggtitle("Time effort by scale")

```




# Richness by effort
```{r, echo=FALSE, warning=FALSE}
# Create a function to generate the plot for each dataset
plot_effort_richness <- function(data, title) {
  ggplot(data, aes(x = effort, y = richness)) +
    geom_point(alpha = 0.5) +  # Set transparency of points to 50%
    geom_smooth(aes(color = "Linear Fit"), method = "lm", se = FALSE, linetype = "dashed") +  
    geom_smooth(aes(color = "Logarithmic Fit"), method = "lm", formula = y ~ log(x), se = FALSE) +  
    geom_smooth(aes(color = "GAM Fit"), method = "gam", formula = y ~ s(x), se = FALSE) + 
    labs(title = title, x = "Effort", y = "Species Richness") +
    scale_color_manual(values = c("Linear Fit" = "red", 
                                    "Logarithmic Fit" = "blue", 
                                    "GAM Fit" = "darkgreen")) +  
    theme_classic() +  # Set classic theme
    theme(
      legend.position = "bottom",
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 14),
      plot.title = element_text(size = 16, face = "bold")
    ) 
}

# Plot for data_1km
plot_1km <- plot_effort_richness(data_1km, "Effort vs. Species Richness (1km)")

# Plot for data_4km
plot_4km <- plot_effort_richness(data_4km, "Effort vs. Species Richness (4km)")

# Plot for data_9km
plot_9km <- plot_effort_richness(data_9km, "Effort vs. Species Richness (9km)")

# Display the plots
plot_1km
plot_4km
plot_9km
```


# Sparing and sharing metrics

## Sharing
```{r, echo=FALSE}
ggplot(combined_data, aes(x = SVI, fill = scale)) +
  geom_density(alpha = 0.5) +
  labs(x = "Sharing", y = "Density") +
  ggtitle("Sharing by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```

## Sparing uneveness
```{r, echo=FALSE}
ggplot(combined_data, aes(x = SEVIo, fill = scale)) +
  geom_density(alpha = 0.5) +
  labs(x = "Sparing uneveness", y = "Density") +
  ggtitle("Sparing uneveness by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```

## Sparing green
```{r, echo=FALSE}
ggplot(combined_data, aes(x = SGVI, fill = scale)) +
  geom_density(alpha = 0.5) +
  labs(x = "Sparing green", y = "Density") +
  ggtitle("Sparing green by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```

## Sparing urban
```{r, echo=FALSE}
ggplot(combined_data, aes(x = SUVI, fill = scale)) +
  geom_density(alpha = 0.5) +
  labs(x = "Sparing urban", y = "Density") +
  ggtitle("Sparing urban by scale") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```

# Scales
## Density
```{r, echo=FALSE}
# Gráficos individuais para 1km
p1_1km <- ggplot(data_1km, aes(x = SVI)) +
  geom_density(alpha = 0.5, fill = "#f7756a", color = "#f7756a") +
  labs(x = "Sharing", y = "Density") +
  ggtitle("Scale 1km²") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  )

p2_1km <- ggplot(data_1km, aes(x = SEVIo)) +
  geom_density(alpha = 0.5, fill = "#f7756a", color = "#f7756a") +
  labs(x = "Sparing uneveness", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p3_1km <- ggplot(data_1km, aes(x = SGVI)) +
  geom_density(alpha = 0.5, fill = "#f7756a", color = "#f7756a") +
  labs(x = "Sparing green", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p4_1km <- ggplot(data_1km, aes(x = SUVI)) +
  geom_density(alpha = 0.5, fill = "#f7756a", color = "#f7756a") +
  labs(x = "Sparing urban", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

# Gráficos individuais para 4km
p1_4km <- ggplot(data_4km, aes(x = SVI)) +
  geom_density(alpha = 0.5, fill = "#60a576", color = "#60a576") +
  labs(x = "Sharing", y = "Density") +
  ggtitle("Scale 4km²") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  )

p2_4km <- ggplot(data_4km, aes(x = SEVIo)) +
  geom_density(alpha = 0.5, fill = "#60a576", color = "#60a576") +
  labs(x = "Sparing uneveness", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p3_4km <- ggplot(data_4km, aes(x = SGVI)) +
  geom_density(alpha = 0.5, fill = "#60a576", color = "#60a576") +
  labs(x = "Sparing green", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p4_4km <- ggplot(data_4km, aes(x = SUVI)) +
  geom_density(alpha = 0.5, fill = "#60a576", color = "#60a576") +
  labs(x = "Sparing urban", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

# Gráficos individuais para 9km
p1_9km <- ggplot(data_9km, aes(x = SVI)) +
  geom_density(alpha = 0.5, fill = "#7b94d2", color = "#7b94d2") +
  labs(x = "Sharing", y = "Density") +
  ggtitle("Scale 9km²") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  )

p2_9km <- ggplot(data_9km, aes(x = SEVIo)) +
  geom_density(alpha = 0.5, fill = "#7b94d2", color = "#7b94d2") +
  labs(x = "Sparing uneveness", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p3_9km <- ggplot(data_9km, aes(x = SGVI)) +
  geom_density(alpha = 0.5, fill = "#7b94d2", color = "#7b94d2") +
  labs(x = "Sparing green", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p4_9km <- ggplot(data_9km, aes(x = SUVI)) +
  geom_density(alpha = 0.5, fill = "#7b94d2", color = "#7b94d2") +
  labs(x = "Sparing urban", y = "Density") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

# Usar patchwork para combinar os gráficos
(p1_1km | p1_4km | p1_9km) / 
(p2_1km | p2_4km | p2_9km) / 
(p3_1km | p3_4km | p3_9km) / 
(p4_1km | p4_4km | p4_9km)
```

## Frequency
```{r, echo=FALSE, warning=FALSE}
# Gráficos individuais para 1km
p1_1km <- ggplot(data_1km, aes(x = SVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#f7756a", color = "#f7756a", bins = 30) +
  labs(x = "Sharing", y = "Frequency") +
  ggtitle("Scale 1km²") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  )

p2_1km <- ggplot(data_1km, aes(x = SEVIo)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#f7756a", color = "#f7756a", bins = 30) +
  labs(x = "Sparing uneveness", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p3_1km <- ggplot(data_1km, aes(x = SGVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#f7756a", color = "#f7756a", bins = 30) +
  labs(x = "Sparing green", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p4_1km <- ggplot(data_1km, aes(x = SUVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#f7756a", color = "#f7756a", bins = 30) +
  labs(x = "Sparing urban", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

# Gráficos individuais para 4km
p1_4km <- ggplot(data_4km, aes(x = SVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#60a576", color = "#60a576", bins = 30) +
  labs(x = "Sharing", y = "Frequency") +
  ggtitle("Scale 4km²") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  )

p2_4km <- ggplot(data_4km, aes(x = SEVIo)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#60a576", color = "#60a576", bins = 30) +
  labs(x = "Sparing uneveness", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p3_4km <- ggplot(data_4km, aes(x = SGVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#60a576", color = "#60a576", bins = 30) +
  labs(x = "Sparing green", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p4_4km <- ggplot(data_4km, aes(x = SUVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#60a576", color = "#60a576", bins = 30) +
  labs(x = "Sparing urban", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

# Gráficos individuais para 9km
p1_9km <- ggplot(data_9km, aes(x = SVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#7b94d2", color = "#7b94d2", bins = 30) +
  labs(x = "Sharing", y = "Frequency") +
  ggtitle("Scale 9km²") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold")
  )

p2_9km <- ggplot(data_9km, aes(x = SEVIo)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#7b94d2", color = "#7b94d2", bins = 30) +
  labs(x = "Sparing uneveness", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p3_9km <- ggplot(data_9km, aes(x = SGVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#7b94d2", color = "#7b94d2", bins = 30) +
  labs(x = "Sparing green", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

p4_9km <- ggplot(data_9km, aes(x = SUVI)) +
  geom_histogram(aes(y = ..count..), alpha = 0.5, fill = "#7b94d2", color = "#7b94d2", bins = 30) +
  labs(x = "Sparing urban", y = "Frequency") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 12)
  )

# Usar patchwork para combinar os gráficos
(p1_1km | p1_4km | p1_9km) / 
(p2_1km | p2_4km | p2_9km) / 
(p3_1km | p3_4km | p3_9km) / 
(p4_1km | p4_4km | p4_9km)
```

## 1 km²
```{r, echo=FALSE}
# Histograma para SVI
p1 <- ggplot(data_4km, aes(x = SVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) + 
  labs(x = "Sharing", y = "Frequency") +
  ggtitle("Sharing (4km²)") +  
  theme_classic() +
  theme(
    legend.position = "none",  
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SEVIo
p2 <- ggplot(data_4km, aes(x = SEVIo, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) + 
  labs(x = "Sparing uneveness", y = "Frequency") +
  ggtitle("Sparing uneveness (4km²)") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SGVI
p3 <- ggplot(data_4km, aes(x = SGVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing green", y = "Frequency") +
  ggtitle("Sparing green (4km²)") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SUVI
p4 <- ggplot(data_4km, aes(x = SUVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing urban", y = "Frequency") +
  ggtitle("Sparing urban (4km²)") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Usar patchwork para combinar os gráficos
(p1 | p2) / (p3 | p4)
```

## 4 km²
```{r, echo=FALSE}
# Histograma para SVI
p1 <- ggplot(data_4km, aes(x = SVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sharing", y = "Frequency") +
  ggtitle("Sharing (4km²)") +
  scale_fill_manual(values = "#60a576") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SEVIo
p2 <- ggplot(data_4km, aes(x = SEVIo, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing uneveness", y = "Frequency") +
  ggtitle("Sparing uneveness (4km²)") +
  scale_fill_manual(values = "#60a576") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SGVI
p3 <- ggplot(data_4km, aes(x = SGVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing green", y = "Frequency") +
  ggtitle("Sparing green (4km²)") +
  scale_fill_manual(values = "#60a576") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SUVI
p4 <- ggplot(data_4km, aes(x = SUVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing urban", y = "Frequency") +
  ggtitle("Sparing urban (4km²)") +
  scale_fill_manual(values = "#60a576") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Usar patchwork para combinar os gráficos
(p1 | p2) / (p3 | p4)
```

## 9 km²
```{r, echo=FALSE}
# Histograma para SVI
p1 <- ggplot(data_9km, aes(x = SVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sharing", y = "Frequency") +
  ggtitle("Sharing (9km²)") +
  scale_fill_manual(values = "#5575c5") + 
  theme_classic() +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SEVIo
p2 <- ggplot(data_9km, aes(x = SEVIo, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing uneveness", y = "Frequency") +
  ggtitle("Sparing uneveness (9km²)") +
  scale_fill_manual(values = "#5575c5") +
  theme_classic() +
  theme(
    legend.position = "none",  # Remove a legenda
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SGVI
p3 <- ggplot(data_9km, aes(x = SGVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing green", y = "Frequency") +
  ggtitle("Sparing green (9km²)") +
  scale_fill_manual(values = "#5575c5") + 
  theme_classic() +
  theme(
    legend.position = "none",  # Remove a legenda
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histograma para SUVI
p4 <- ggplot(data_9km, aes(x = SUVI, fill = scale)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(x = "Sparing urban", y = "Frequency") +
  ggtitle("Sparing urban (9km²)") +
  scale_fill_manual(values = "#5575c5") +
  theme_classic() +
  theme(
    legend.position = "none",  # Remove a legenda
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Usar patchwork para combinar os gráficos
(p1 | p2) / (p3 | p4)
```